### 2024-07-24 paper.qmd
initial quarto files commit

---
title: "Cloud-Optimized HDF5 for NASA’s ICESat-2 Mission"
format:
    agu-pdf:
        keep-tex: true
    agu-html: default
author:
  - name: Luis A. Lopez
    affiliations:
      - name: National Snow and Ice Data Center, University of Colorado, Boulder.
        department: CIRES
        address: CIRES, 449 UCB
        city: Boulder
        region: CO
        country: USA
        postal-code: 80309
    orcid: 0000-0003-4896-3263
    email: luis.lopezespinosa@colorado.edu
    url: https://github.com/betolink
  - name: "Andrew P. Barrett"
    affiliations:
      - name: National Snow and Ice Data Center, University of Colorado, Boulder.
        department: CIRES
        address: CIRES, 449 UCB
        city: Boulder
        region: CO
        country: USA
        postal-code: 80309
    orcid: 0000-0003-4394-5445
    url: "https://github.com/andypbarrett"
  - name: "Amy Steiker"
    affiliations:
      - name: National Snow and Ice Data Center, University of Colorado, Boulder.
        department: CIRES
        address: CIRES, 449 UCB
        city: Boulder
        region: CO
        country: USA
        postal-code: 80309
    orcid: 0000-0002-3039-0260
    url: "https://github.com/asteiker"

abstract: |
  The Hierarchical Data Format (HDF) is a common archival format for n-dimensional scientific data; it has been utilized to store valuable information from astrophysics to earth sciences and everything in between. As flexible and powerful as HDF can be, it comes with big tradeoffs when it’s accessed from remote storage systems, mainly because the file format and the client I/O libraries were designed for local and supercomputing workflows. As scientific data and workflows migrate to the cloud , efficient access to data stored in HDF format is a key factor that will accelerate or slow down “science in the cloud” across all disciplines.
  We present an implementation of recently available features in the HDF5 stack that results in performant access to HDF from remote cloud storage. This performance is on par with modern cloud-native formats like Zarr but with the advantage of not having to reformat data or generate metadata sidecar files (DMR++, Kerchunk). Our benchmarks also show potential cost-savings for  data producers if their data are processed using cloud-optimized strategies.

keywords: ["cloud-native","cloud", "HDF5", "NASA", "ICESat-2"]
key-points:
  - Key Points convey the main points and conclusions of the article. 
  - Up to three key point statements are allowed, and each is limited to at most 140 characters with no abbreviations.
  - Key Points are not included in the word count.
bibliography: bibliography.bib  
citation:
  container-title: Geophysical Research Letters
keep-tex: true
date: last-modified
---

## Section Heading

Lorem ipsum dolor sit amet, consectetur adipiscing elit. Vestibulum hendrerit facilisis velit sit amet malesuada. Phasellus ornare nibh augue, maximus sodales ex tristique vitae. Vivamus non sollicitudin orci, aliquam placerat metus. Maecenas volutpat orci felis, vel finibus urna consectetur sed. Integer in dui ac dui mollis imperdiet. Quisque sed dapibus nibh. Aenean non luctus leo. Phasellus luctus mauris id aliquet dictum. Aliquam fermentum semper massa, vel dignissim nibh dictum et. See @Hubbard2021.

Phasellus interdum tincidunt ex, a euismod massa pulvinar at. Ut fringilla ut nisi nec volutpat. Morbi imperdiet congue tincidunt. Vivamus eget rutrum purus. Etiam et pretium justo. Donec et egestas sem. Donec molestie ex sit amet viverra egestas. Nullam justo nulla, fringilla at iaculis in, posuere non mauris. Ut eget imperdiet elit.

In luctus mauris vitae imperdiet luctus. Morbi volutpat ligula ut tortor fermentum, eu ornare felis luctus. Donec semper diam vitae mattis posuere. Suspendisse facilisis purus nisi, sit amet egestas ex tempor ut. Cras tortor nulla, euismod at fermentum vel, dictum vel justo. Aenean commodo interdum diam nec placerat. Nunc vestibulum felis at est tincidunt, at euismod dui vestibulum. Nulla venenatis tortor at auctor iaculis. Donec consectetur neque ut sagittis ornare. Nullam pharetra felis tempor suscipit efficitur. Curabitur nibh ex, euismod at congue hendrerit, egestas id mi. Duis porttitor neque in commodo elementum. Fusce vitae fermentum nisi, euismod viverra augue. Curabitur at mi pretium, accumsan purus nec, tempus turpis.

Donec non semper dui, quis aliquet est. Quisque quis sapien at massa ultricies egestas. Duis consequat ultricies erat, a pulvinar nisl vestibulum id. Sed tristique turpis ligula, et tempor lectus iaculis at. Vivamus commodo sapien ac turpis vestibulum dapibus. Morbi tristique arcu metus, et laoreet nisi varius nec. Pellentesque habitant morbi tristique senectus et netus et malesuada fames ac turpis egestas. Fusce sit amet nisl at mauris suscipit aliquet. Nulla vitae dignissim urna. Suspendisse sit amet arcu vitae magna blandit mattis. Vivamus convallis efficitur pulvinar. Sed cursus elit nulla. Sed porta, arcu a euismod pretium, odio dui lacinia lacus, ac vulputate nulla augue eget ex. Nullam consequat ligula sit amet mattis aliquam. Nulla risus urna, ultrices vel ullamcorper id, ornare viverra nunc.

Nunc in lobortis lacus. Duis maximus urna leo, varius sodales arcu interdum nec. Pellentesque imperdiet dolor in leo eleifend dapibus. Ut dapibus, lectus non viverra gravida, ipsum ex faucibus tellus, quis iaculis risus tellus eget augue. Nullam a viverra est. Cras velit nisi, interdum in lacus at, vehicula mattis elit. Curabitur eu viverra purus. Proin pellentesque, metus vitae congue convallis, lorem metus feugiat mi, sit amet auctor purus ligula bibendum ante. Nam id justo scelerisque, rhoncus lectus in, fermentum libero. Donec tincidunt egestas ex ac eleifend. Cras faucibus ipsum a nunc faucibus fermentum. Integer et maximus lacus. Nam dictum nibh id viverra convallis.

## Acknowledgments

Phasellus interdum tincidunt ex, a euismod massa pulvinar at. Ut fringilla ut nisi nec volutpat. Morbi imperdiet congue tincidunt. Vivamus eget rutrum purus. Etiam et pretium justo. Donec et egestas sem. Donec molestie ex sit amet viverra egestas. Nullam justo nulla, fringilla at iaculis in, posuere non mauris. Ut eget imperdiet elit.

## Open research

Phasellus interdum tincidunt ex, a euismod massa pulvinar at. Ut fringilla ut nisi nec volutpat. Morbi imperdiet congue tincidunt. Vivamus eget rutrum purus. Etiam et pretium justo. Donec et egestas sem. Donec molestie ex sit amet viverra egestas. Nullam justo nulla, fringilla at iaculis in, posuere non mauris. Ut eget imperdiet elit.

## References {.unnumbered}

:::{#refs}

:::

### 2024-07-24 _tex/paper.tex
Built site for gh-pages

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{agujournal2019}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{url} %this package should fix any errors with URLs in refs.
\usepackage{lineno}
\usepackage[inline]{trackchanges} %for better track changes. finalnew option will compile document with changes incorporated.
\usepackage{soul}
\linenumbers
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Cloud-Optimized HDF5 for NASA's ICESat-2 Mission},
  pdfauthor={Luis A. Lopez; Andrew P. Barrett; Amy Steiker},
  pdfkeywords={cloud-native, cloud, HDF5, NASA, ICESat-2},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\journalname{Geophysical Research Letters}

\draftfalse

\begin{document}
\title{Cloud-Optimized HDF5 for NASA's ICESat-2 Mission}

\authors{Luis A. Lopez\affil{1}, Andrew P. Barrett\affil{1}, Amy
Steiker\affil{1}}
\affiliation{1}{CIRES, National Snow and Ice Data Center, University of
Colorado, Boulder., Boulder, CO, USA}

\begin{abstract}
The Hierarchical Data Format (HDF) is a common archival format for
n-dimensional scientific data; it has been utilized to store valuable
information from astrophysics to earth sciences and everything in
between. As flexible and powerful as HDF can be, it comes with big
tradeoffs when it's accessed from remote storage systems, mainly because
the file format and the client I/O libraries were designed for local and
supercomputing workflows. As scientific data and workflows migrate to
the cloud , efficient access to data stored in HDF format is a key
factor that will accelerate or slow down ``science in the cloud'' across
all disciplines. We present an implementation of recently available
features in the HDF5 stack that results in performant access to HDF from
remote cloud storage. This performance is on par with modern
cloud-native formats like Zarr but with the advantage of not having to
reformat data or generate metadata sidecar files (DMR++, Kerchunk). Our
benchmarks also show potential cost-savings for data producers if their
data are processed using cloud-optimized strategies.
\end{abstract}

\section{Section Heading}\label{section-heading}

Lorem ipsum dolor sit amet, consectetur adipiscing elit. Vestibulum
hendrerit facilisis velit sit amet malesuada. Phasellus ornare nibh
augue, maximus sodales ex tristique vitae. Vivamus non sollicitudin
orci, aliquam placerat metus. Maecenas volutpat orci felis, vel finibus
urna consectetur sed. Integer in dui ac dui mollis imperdiet. Quisque
sed dapibus nibh. Aenean non luctus leo. Phasellus luctus mauris id
aliquet dictum. Aliquam fermentum semper massa, vel dignissim nibh
dictum et. See Hubbard et al. (2021).

Phasellus interdum tincidunt ex, a euismod massa pulvinar at. Ut
fringilla ut nisi nec volutpat. Morbi imperdiet congue tincidunt.
Vivamus eget rutrum purus. Etiam et pretium justo. Donec et egestas sem.
Donec molestie ex sit amet viverra egestas. Nullam justo nulla,
fringilla at iaculis in, posuere non mauris. Ut eget imperdiet elit.

In luctus mauris vitae imperdiet luctus. Morbi volutpat ligula ut tortor
fermentum, eu ornare felis luctus. Donec semper diam vitae mattis
posuere. Suspendisse facilisis purus nisi, sit amet egestas ex tempor
ut. Cras tortor nulla, euismod at fermentum vel, dictum vel justo.
Aenean commodo interdum diam nec placerat. Nunc vestibulum felis at est
tincidunt, at euismod dui vestibulum. Nulla venenatis tortor at auctor
iaculis. Donec consectetur neque ut sagittis ornare. Nullam pharetra
felis tempor suscipit efficitur. Curabitur nibh ex, euismod at congue
hendrerit, egestas id mi. Duis porttitor neque in commodo elementum.
Fusce vitae fermentum nisi, euismod viverra augue. Curabitur at mi
pretium, accumsan purus nec, tempus turpis.

Donec non semper dui, quis aliquet est. Quisque quis sapien at massa
ultricies egestas. Duis consequat ultricies erat, a pulvinar nisl
vestibulum id. Sed tristique turpis ligula, et tempor lectus iaculis at.
Vivamus commodo sapien ac turpis vestibulum dapibus. Morbi tristique
arcu metus, et laoreet nisi varius nec. Pellentesque habitant morbi
tristique senectus et netus et malesuada fames ac turpis egestas. Fusce
sit amet nisl at mauris suscipit aliquet. Nulla vitae dignissim urna.
Suspendisse sit amet arcu vitae magna blandit mattis. Vivamus convallis
efficitur pulvinar. Sed cursus elit nulla. Sed porta, arcu a euismod
pretium, odio dui lacinia lacus, ac vulputate nulla augue eget ex.
Nullam consequat ligula sit amet mattis aliquam. Nulla risus urna,
ultrices vel ullamcorper id, ornare viverra nunc.

Nunc in lobortis lacus. Duis maximus urna leo, varius sodales arcu
interdum nec. Pellentesque imperdiet dolor in leo eleifend dapibus. Ut
dapibus, lectus non viverra gravida, ipsum ex faucibus tellus, quis
iaculis risus tellus eget augue. Nullam a viverra est. Cras velit nisi,
interdum in lacus at, vehicula mattis elit. Curabitur eu viverra purus.
Proin pellentesque, metus vitae congue convallis, lorem metus feugiat
mi, sit amet auctor purus ligula bibendum ante. Nam id justo
scelerisque, rhoncus lectus in, fermentum libero. Donec tincidunt
egestas ex ac eleifend. Cras faucibus ipsum a nunc faucibus fermentum.
Integer et maximus lacus. Nam dictum nibh id viverra convallis.

\section{Acknowledgments}\label{acknowledgments}

Phasellus interdum tincidunt ex, a euismod massa pulvinar at. Ut
fringilla ut nisi nec volutpat. Morbi imperdiet congue tincidunt.
Vivamus eget rutrum purus. Etiam et pretium justo. Donec et egestas sem.
Donec molestie ex sit amet viverra egestas. Nullam justo nulla,
fringilla at iaculis in, posuere non mauris. Ut eget imperdiet elit.

\section{Open research}\label{open-research}

Phasellus interdum tincidunt ex, a euismod massa pulvinar at. Ut
fringilla ut nisi nec volutpat. Morbi imperdiet congue tincidunt.
Vivamus eget rutrum purus. Etiam et pretium justo. Donec et egestas sem.
Donec molestie ex sit amet viverra egestas. Nullam justo nulla,
fringilla at iaculis in, posuere non mauris. Ut eget imperdiet elit.

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\vspace{1em}

\bibitem[\citeproctext]{ref-Hubbard2021}
Hubbard, B., Christoffersen, P., Doyle, S. H., Chudley, T. R.,
Schoonman, C. M., Law, R., \& Bougamont, M. (2021). Borehole-based
characterization of deep mixed-mode crevasses at a greenlandic outlet
glacier. \emph{{AGU} Advances}, \emph{2}(2).
\url{https://doi.org/10.1029/2020av000291}

\end{CSLReferences}

\end{document}

### 2024-08-19 paper.qmd
update paper text, still work in progress

@@
 ---
-title: "Cloud-Optimized HDF5 for NASA’s ICESat-2 Mission"
+title: "Evaluating Cloud-Optimized HDF5 for NASA’s ICESat-2 Mission"
 format:
@@
     url: "https://github.com/asteiker"
+  - name: "Aleksandar Jelenak"
+    affiliations:
+      - name: The HDF Group
+        city: Champaign
+        region: IL
+        country: USA
+        postal-code: 61820
+    orcid: 0009-0001-2102-0559
+    url: "https://github.com/ajelenak"
+  - name: "Lisa Kaser"
+    affiliations:
+      - name: National Snow and Ice Data Center, University of Colorado, Boulder.
+        department: CIRES
+        address: CIRES, 449 UCB
+        city: Boulder
+        region: CO
+        country: USA
+        postal-code: 80309
+    url: "https://nsidc.org/about/about-nsidc/what-we-do/our-people/lisa_kaser"
+  - name: "Jeffrey E. Lee"
+    affiliations:
+      - name:  NASA / KBR
+        department: NASA Goddard Space Flight Center 
+        address: 8800 Greenbelt Rd
+        city: Greenbelt
+        region: MD
+        country: USA
+        postal-code: 20771
+    url: "https://science.gsfc.nasa.gov/sci/bio/jeffrey.e.lee"
 
@@
 
-## Section Heading
+## Problem
 
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Vestibulum hendrerit facilisis velit sit amet malesuada.
-Phasellus ornare nibh augue, maximus sodales ex tristique vitae.
-Vivamus non sollicitudin orci, aliquam placerat metus.
-Maecenas volutpat orci felis, vel finibus urna consectetur sed.
-Integer in dui ac dui mollis imperdiet.
-Quisque sed dapibus nibh.
-Aenean non luctus leo.
-Phasellus luctus mauris id aliquet dictum.
-Aliquam fermentum semper massa, vel dignissim nibh dictum et.
-See @Hubbard2021.
+Scientific data from NASA and other agencies are increasingly being distributed from the commercial cloud.
+Cloud storage enables large-scale workflows and should reduce local storage costs.
+It also allows the use of scalable on-demand cloud computing resources by individual scientists and the broader scientific community.
+However, the majority of this scientific data is stored in a format that was not designed for the cloud: The Hierarchical Data format or HDF.
 
-Phasellus interdum tincidunt ex, a euismod massa pulvinar at.
-Ut fringilla ut nisi nec volutpat.
-Morbi imperdiet congue tincidunt.
-Vivamus eget rutrum purus.
-Etiam et pretium justo.
-Donec et egestas sem.
-Donec molestie ex sit amet viverra egestas.
-Nullam justo nulla, fringilla at iaculis in, posuere non mauris.
-Ut eget imperdiet elit.
+The most recent version of the Hierarchical Data Format is HDF5, a common archival format for n-dimensional scientific data; it has been utilized to store valuable information from astrophysics to earth sciences and everything in between.
+As flexible and powerful as HDF5 can be, it comes with big tradeoffs when it’s accessed from remote storage systems.
 
-In luctus mauris vitae imperdiet luctus.
-Morbi volutpat ligula ut tortor fermentum, eu ornare felis luctus.
-Donec semper diam vitae mattis posuere.
-Suspendisse facilisis purus nisi, sit amet egestas ex tempor ut.
-Cras tortor nulla, euismod at fermentum vel, dictum vel justo.
-Aenean commodo interdum diam nec placerat.
-Nunc vestibulum felis at est tincidunt, at euismod dui vestibulum.
-Nulla venenatis tortor at auctor iaculis.
-Donec consectetur neque ut sagittis ornare.
-Nullam pharetra felis tempor suscipit efficitur.
-Curabitur nibh ex, euismod at congue hendrerit, egestas id mi.
-Duis porttitor neque in commodo elementum.
-Fusce vitae fermentum nisi, euismod viverra augue.
-Curabitur at mi pretium, accumsan purus nec, tempus turpis.
+HDF5 is a complex file format; we can think of it as a file system using a tree-like structure with multiple data types and native data structures.
+Because of this complexity, the most reliable way of accessing data stored in this format is using the HDF5 C API.
+Regardless of access pattern, nearly all tools ultimately rely on the HDF5-C library and this brings a couple issues that affect the efficiency of accessing this format over the network:
 
-Donec non semper dui, quis aliquet est.
-Quisque quis sapien at massa ultricies egestas.
-Duis consequat ultricies erat, a pulvinar nisl vestibulum id.
-Sed tristique turpis ligula, et tempor lectus iaculis at.
-Vivamus commodo sapien ac turpis vestibulum dapibus.
-Morbi tristique arcu metus, et laoreet nisi varius nec.
-Pellentesque habitant morbi tristique senectus et netus et malesuada fames ac turpis egestas.
-Fusce sit amet nisl at mauris suscipit aliquet.
-Nulla vitae dignissim urna.
-Suspendisse sit amet arcu vitae magna blandit mattis.
-Vivamus convallis efficitur pulvinar.
-Sed cursus elit nulla.
-Sed porta, arcu a euismod pretium, odio dui lacinia lacus, ac vulputate nulla augue eget ex.
-Nullam consequat ligula sit amet mattis aliquam.
-Nulla risus urna, ultrices vel ullamcorper id, ornare viverra nunc.
+---
 
-Nunc in lobortis lacus.
-Duis maximus urna leo, varius sodales arcu interdum nec.
-Pellentesque imperdiet dolor in leo eleifend dapibus.
-Ut dapibus, lectus non viverra gravida, ipsum ex faucibus tellus, quis iaculis risus tellus eget augue.
-Nullam a viverra est.
-Cras velit nisi, interdum in lacus at, vehicula mattis elit.
-Curabitur eu viverra purus.
-Proin pellentesque, metus vitae congue convallis, lorem metus feugiat mi, sit amet auctor purus ligula bibendum ante.
-Nam id justo scelerisque, rhoncus lectus in, fermentum libero.
-Donec tincidunt egestas ex ac eleifend.
-Cras faucibus ipsum a nunc faucibus fermentum.
-Integer et maximus lacus.
-Nam dictum nibh id viverra convallis.
+#### **Metadata fragmentation**
 
-## Acknowledgments
+By default, file-level metadata associated with a dataset is stored in chunks of 4kb. This produces a lot of fragmentation across the file especially for data with many variables and nested groups.
 
-Phasellus interdum tincidunt ex, a euismod massa pulvinar at.
-Ut fringilla ut nisi nec volutpat.
-Morbi imperdiet congue tincidunt.
-Vivamus eget rutrum purus.
-Etiam et pretium justo.
-Donec et egestas sem.
-Donec molestie ex sit amet viverra egestas.
-Nullam justo nulla, fringilla at iaculis in, posuere non mauris.
-Ut eget imperdiet elit.
+#### **Global API Lock**
 
-## Open research
+Because of the historical complexity of operations with the HDF5 format, there has been a necessity to make the library thread-safe and similarly to what happens in the Python language, the simplest mechanism to implement this is to have a global API lock.
+This global lock is not as big of an issue when we read data from local disk but it becomes a major bottleneck when we read data over the network because each read is sequential and latency in the cloud is exponentially bigger than local access.
 
-Phasellus interdum tincidunt ex, a euismod massa pulvinar at.
-Ut fringilla ut nisi nec volutpat.
-Morbi imperdiet congue tincidunt.
-Vivamus eget rutrum purus.
-Etiam et pretium justo.
-Donec et egestas sem.
-Donec molestie ex sit amet viverra egestas.
-Nullam justo nulla, fringilla at iaculis in, posuere non mauris.
-Ut eget imperdiet elit.
+---
+
+::: {#fig-1 fig-env="figure*"}
+
+![](figures/figure-1.png)
+
+shows how reads (Rn) are done in order to access file metadata, In the first read, R0, the HDF5 library verifies the file signature from the superblock,  subsequent reads, R1, R2,...Rn, read file metadata, 4kb at the time.
+
+:::
+
+#### Background and data selection
+
+As a result of community feedback and “hack weeks” organized by NSIDC and UW eScience Institute in 2023, NSIDC started the Cloud Optimized Format Investigation (COFI) project to improve access to HDF5 from the ICESat-2 mission.
+A spaceborne lidar that retrieves surface topography of the Earth’s ice sheets, land and [oceans @NEUMANN2019111325].
+Because of its complexity, large size and importance for cryospheric studies we targeted the ATL03 dataset.
+ATL03 core data are geolocated photon heights from the ICESat-2 ATLAS instrument.
+Each file contains 1003 geophysical variables in 6 data groups.
+Although our research was focused on this dataset, most of our findings are applicable to any dataset stored in HDF5 and NetCDF4.
+
+## Methodology
+
+We tested access times to original and cloud-optimized small (1 GB), medium (2 GB) and large (7 GB) HDF5 ATL03 files [list files tested] stored in AWS S3 buckets in region us-west-2, the region hosting NASA’s Earthdata Cloud archives.
+Files were accessed using Python tools commonly used by Earth scientists: h5py and Xarray.
+h5py is a Python wrapper around the HDF5 C API.
+xarray^[`h5py` is a dependency of Xarray]  is a widely used Python package for working with n-dimensional data.
+We also tested access times using h5coro, a python package optimized for reading HDF5 files from S3 buckets and kerchunk, a tool that creates an efficient lookup table for file chunks to allow performant partial reads of files.
+
+HDF5 ATL03 files were originally cloud optimized by  “repacking” them, using a relatively new feature in the HDF5 C API called “paged aggregation”.
+Page aggregation does 2 things: it collects file-level metadata from datasets and stores it on dedicated metadata blocks in the file; and it forces the library to write data using fixed-size blocks.
+Aggregation allows client libraries to read  file metadata with only a few requests and uses the page size used in the aggregation as the minimal request size, overriding the 1 request per chunk behavior.
+
+::: {#fig-2 fig-env="figure*"}
+
+![](figures/figure-2.png)
+
+shows how file-level metadata and data gets internally packed once we use paged aggregation on a file.
+
+:::
+
+## Results
+
+::: {#fig-3 fig-env="figure*"}
+
+![](figures/figure-3.png)
+
+Benchmarks show that cloud optimizing ATL03 files improved access times at least an order of magnitude when used with aligned I/O patterns, this is telling the library about the cloud optimization and page size.
+
+:::
+
+## Recommendations
+
+We have split our recommendations for the ATL03 product into 3 main categories, creating the files, accessing the files, and future tool development.
+
+### Recommended cloud optimizations
+
+Based on our testing we recommend the following cloud optimizations for creating HDF5 files for the ATL03 product:
+Create HDF5 files using paged aggregation by setting HDF5 library parameters:
+
+1. File page strategy: H5F_FSPACE_STRATEGY_PAGE
+2. File page size: 8000000
+  If repacking an existing file, h5repack contains the code to alter tese variables inside the file 
+  ```bash
+   h5repack -S PAGE -G 8000000 input.h5 output.h5
+  ```
+3. Avoid using unlimited dimensions when creating variables because the HDF5 API cannot support it inside bffered pages and representation of these variables is not supported by Kerchunk.
+
+#### Reasoning
+
+Based on the variable size of ATL03 it becomes really difficult to allocate a fixed metadata page, big files contain north of 30MB of metadata, but the median sized file is below 8MB. If we had
+adopted user block we would have caused an increase in the file size and storage cost of approximate 30% (reference to our tests).
+Another consequence of using a dedicated fixed page for file-level metadata is
+that metadata overflow will generate the same impact in access times, the library will fetch the metadata in one go but the rest will be using the predefined block size of 4kb.
+
+Paged aggregation is thus the simplest way of cloud optimizing an HDF5 file as the metadata will keep filling dedicated pages until all the file-level metadata is stored at the front of the file.
+Chunk sizes cannot be larger than the page size and when chunk sizes are smaller we need to take into account how these chunks will fit on a page, in an ideal scenario all the space
+will be filled but that is not the case and we will end up with unused space [See @fig-2].
+
+### Recommended access patterns
+
+Placeholder
+
+### Recommended tooling development
+
+Placeholder
+
+### Mission implementation
+
+ATL03 is a complex science data product containing both segmented (20 meters along-track) and large, variable-rate photon datasets.
+ATL03 is created using pipeline-style processing where the science data and NetCDF-style metadata are written by independent software packages.
+The following steps were employed to create cloud-optimized Release 007 ATL03 products, while minimizing increases in file size:
+
+1. Set the "file space strategy" to H5F_FSPACE_STRATEGY_PAGE and enabled "free space tracking" within the HDF5 file creation property list.
+2. Set the "file space page size" to 8MiB.
+3. Changed all "COMPACT" dataset storage types to "CONTIGUOUS".
+4. Increased the "chunk size" of the photon-rate datasets (from 10,000 to 100,000 elements), while making sure no "chunk sizes" exceeded the 8MiB "file space page size".
+5. Introduced a new production step that executes the "h5repack" utility (with no options) to create a "defragmented" final product.
+
+## Discussion
+
+1. Chunking shapes and sizes
+2. Paged aggregation vs User block
+3. Side effects on different access patterns, e.g. Kerchunk
+
+<!-- ## Acknowledgments -->
 

### 2024-08-19 _tex/paper.tex
Built site for gh-pages

@@
 \makeatother
+
+\usepackage{color}
+\usepackage{fancyvrb}
+\newcommand{\VerbBar}{|}
+\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
+\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
+% Add ',fontsize=\small' for more characters per line
+\usepackage{framed}
+\definecolor{shadecolor}{RGB}{241,243,245}
+\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
+\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
+\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.40,0.45,0.13}{#1}}
+\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\BuiltInTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\CharTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
+\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
+\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
+\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
+\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{\textbf{#1}}}
+\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
+\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\ExtensionTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.28,0.35,0.67}{#1}}
+\newcommand{\ImportTok}[1]{\textcolor[rgb]{0.00,0.46,0.62}{#1}}
+\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
+\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{\textbf{#1}}}
+\newcommand{\NormalTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
+\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\RegionMarkerTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
+\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
+\newcommand{\StringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
+\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.07,0.07,0.07}{#1}}
+\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
+\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
 
@@
 \hypersetup{
-  pdftitle={Cloud-Optimized HDF5 for NASA's ICESat-2 Mission},
-  pdfauthor={Luis A. Lopez; Andrew P. Barrett; Amy Steiker},
+  pdftitle={Evaluating Cloud-Optimized HDF5 for NASA's ICESat-2 Mission},
+  pdfauthor={Luis A. Lopez; Andrew P. Barrett; Amy Steiker; Aleksandar Jelenak; Lisa Kaser; Jeffrey E. Lee},
   pdfkeywords={cloud-native, cloud, HDF5, NASA, ICESat-2},
@@
 \begin{document}
-\title{Cloud-Optimized HDF5 for NASA's ICESat-2 Mission}
+\title{Evaluating Cloud-Optimized HDF5 for NASA's ICESat-2 Mission}
 
 \authors{Luis A. Lopez\affil{1}, Andrew P. Barrett\affil{1}, Amy
-Steiker\affil{1}}
+Steiker\affil{1}, Aleksandar Jelenak\affil{2}, Lisa
+Kaser\affil{1}, Jeffrey E. Lee\affil{3}}
 \affiliation{1}{CIRES, National Snow and Ice Data Center, University of
-Colorado, Boulder., Boulder, CO, USA}
+Colorado, Boulder., Boulder, CO, USA}\affiliation{2}{The HDF
+Group, Champaign, IL, USA}\affiliation{3}{NASA Goddard Space Flight
+Center, NASA / KBR, Greenbelt, MD, USA}
 
@@
 
-\section{Section Heading}\label{section-heading}
+\section{Problem}\label{problem}
 
-Lorem ipsum dolor sit amet, consectetur adipiscing elit. Vestibulum
-hendrerit facilisis velit sit amet malesuada. Phasellus ornare nibh
-augue, maximus sodales ex tristique vitae. Vivamus non sollicitudin
-orci, aliquam placerat metus. Maecenas volutpat orci felis, vel finibus
-urna consectetur sed. Integer in dui ac dui mollis imperdiet. Quisque
-sed dapibus nibh. Aenean non luctus leo. Phasellus luctus mauris id
-aliquet dictum. Aliquam fermentum semper massa, vel dignissim nibh
-dictum et. See Hubbard et al. (2021).
+Scientific data from NASA and other agencies are increasingly being
+distributed from the commercial cloud. Cloud storage enables large-scale
+workflows and should reduce local storage costs. It also allows the use
+of scalable on-demand cloud computing resources by individual scientists
+and the broader scientific community. However, the majority of this
+scientific data is stored in a format that was not designed for the
+cloud: The Hierarchical Data format or HDF.
 
-Phasellus interdum tincidunt ex, a euismod massa pulvinar at. Ut
-fringilla ut nisi nec volutpat. Morbi imperdiet congue tincidunt.
-Vivamus eget rutrum purus. Etiam et pretium justo. Donec et egestas sem.
-Donec molestie ex sit amet viverra egestas. Nullam justo nulla,
-fringilla at iaculis in, posuere non mauris. Ut eget imperdiet elit.
+The most recent version of the Hierarchical Data Format is HDF5, a
+common archival format for n-dimensional scientific data; it has been
+utilized to store valuable information from astrophysics to earth
+sciences and everything in between. As flexible and powerful as HDF5 can
+be, it comes with big tradeoffs when it's accessed from remote storage
+systems.
 
-In luctus mauris vitae imperdiet luctus. Morbi volutpat ligula ut tortor
-fermentum, eu ornare felis luctus. Donec semper diam vitae mattis
-posuere. Suspendisse facilisis purus nisi, sit amet egestas ex tempor
-ut. Cras tortor nulla, euismod at fermentum vel, dictum vel justo.
-Aenean commodo interdum diam nec placerat. Nunc vestibulum felis at est
-tincidunt, at euismod dui vestibulum. Nulla venenatis tortor at auctor
-iaculis. Donec consectetur neque ut sagittis ornare. Nullam pharetra
-felis tempor suscipit efficitur. Curabitur nibh ex, euismod at congue
-hendrerit, egestas id mi. Duis porttitor neque in commodo elementum.
-Fusce vitae fermentum nisi, euismod viverra augue. Curabitur at mi
-pretium, accumsan purus nec, tempus turpis.
+HDF5 is a complex file format; we can think of it as a file system using
+a tree-like structure with multiple data types and native data
+structures. Because of this complexity, the most reliable way of
+accessing data stored in this format is using the HDF5 C API. Regardless
+of access pattern, nearly all tools ultimately rely on the HDF5-C
+library and this brings a couple issues that affect the efficiency of
+accessing this format over the network:
 
-Donec non semper dui, quis aliquet est. Quisque quis sapien at massa
-ultricies egestas. Duis consequat ultricies erat, a pulvinar nisl
-vestibulum id. Sed tristique turpis ligula, et tempor lectus iaculis at.
-Vivamus commodo sapien ac turpis vestibulum dapibus. Morbi tristique
-arcu metus, et laoreet nisi varius nec. Pellentesque habitant morbi
-tristique senectus et netus et malesuada fames ac turpis egestas. Fusce
-sit amet nisl at mauris suscipit aliquet. Nulla vitae dignissim urna.
-Suspendisse sit amet arcu vitae magna blandit mattis. Vivamus convallis
-efficitur pulvinar. Sed cursus elit nulla. Sed porta, arcu a euismod
-pretium, odio dui lacinia lacus, ac vulputate nulla augue eget ex.
-Nullam consequat ligula sit amet mattis aliquam. Nulla risus urna,
-ultrices vel ullamcorper id, ornare viverra nunc.
+\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}
 
-Nunc in lobortis lacus. Duis maximus urna leo, varius sodales arcu
-interdum nec. Pellentesque imperdiet dolor in leo eleifend dapibus. Ut
-dapibus, lectus non viverra gravida, ipsum ex faucibus tellus, quis
-iaculis risus tellus eget augue. Nullam a viverra est. Cras velit nisi,
-interdum in lacus at, vehicula mattis elit. Curabitur eu viverra purus.
-Proin pellentesque, metus vitae congue convallis, lorem metus feugiat
-mi, sit amet auctor purus ligula bibendum ante. Nam id justo
-scelerisque, rhoncus lectus in, fermentum libero. Donec tincidunt
-egestas ex ac eleifend. Cras faucibus ipsum a nunc faucibus fermentum.
-Integer et maximus lacus. Nam dictum nibh id viverra convallis.
+\subsubsection{\texorpdfstring{\textbf{Metadata
+fragmentation}}{Metadata fragmentation}}\label{metadata-fragmentation}
 
-\section{Acknowledgments}\label{acknowledgments}
+By default, file-level metadata associated with a dataset is stored in
+chunks of 4kb. This produces a lot of fragmentation across the file
+especially for data with many variables and nested groups.
 
-Phasellus interdum tincidunt ex, a euismod massa pulvinar at. Ut
-fringilla ut nisi nec volutpat. Morbi imperdiet congue tincidunt.
-Vivamus eget rutrum purus. Etiam et pretium justo. Donec et egestas sem.
-Donec molestie ex sit amet viverra egestas. Nullam justo nulla,
-fringilla at iaculis in, posuere non mauris. Ut eget imperdiet elit.
+\subsubsection{\texorpdfstring{\textbf{Global API
+Lock}}{Global API Lock}}\label{global-api-lock}
 
-\section{Open research}\label{open-research}
+Because of the historical complexity of operations with the HDF5 format,
+there has been a necessity to make the library thread-safe and similarly
+to what happens in the Python language, the simplest mechanism to
+implement this is to have a global API lock. This global lock is not as
+big of an issue when we read data from local disk but it becomes a major
+bottleneck when we read data over the network because each read is
+sequential and latency in the cloud is exponentially bigger than local
+access.
 
-Phasellus interdum tincidunt ex, a euismod massa pulvinar at. Ut
-fringilla ut nisi nec volutpat. Morbi imperdiet congue tincidunt.
-Vivamus eget rutrum purus. Etiam et pretium justo. Donec et egestas sem.
-Donec molestie ex sit amet viverra egestas. Nullam justo nulla,
-fringilla at iaculis in, posuere non mauris. Ut eget imperdiet elit.
+\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}
+
+\begin{figure*}
+
+\centering{
+
+\includegraphics{figures/figure-1.png}
+
+}
+
+\caption{\label{fig-1}shows how reads (Rn) are done in order to access
+file metadata, In the first read, R0, the HDF5 library verifies the file
+signature from the superblock, subsequent reads, R1, R2,\ldots Rn, read
+file metadata, 4kb at the time.}
+
+\end{figure*}%
+
+\subsubsection{Background and data
+selection}\label{background-and-data-selection}
+
+As a result of community feedback and ``hack weeks'' organized by NSIDC
+and UW eScience Institute in 2023, NSIDC started the Cloud Optimized
+Format Investigation (COFI) project to improve access to HDF5 from the
+ICESat-2 mission. A spaceborne lidar that retrieves surface topography
+of the Earth's ice sheets, land and (oceans Neumann et al., 2019).
+Because of its complexity, large size and importance for cryospheric
+studies we targeted the ATL03 dataset. ATL03 core data are geolocated
+photon heights from the ICESat-2 ATLAS instrument. Each file contains
+1003 geophysical variables in 6 data groups. Although our research was
+focused on this dataset, most of our findings are applicable to any
+dataset stored in HDF5 and NetCDF4.
+
+\section{Methodology}\label{methodology}
+
+We tested access times to original and cloud-optimized small (1 GB),
+medium (2 GB) and large (7 GB) HDF5 ATL03 files {[}list files tested{]}
+stored in AWS S3 buckets in region us-west-2, the region hosting NASA's
+Earthdata Cloud archives. Files were accessed using Python tools
+commonly used by Earth scientists: h5py and Xarray. h5py is a Python
+wrapper around the HDF5 C API. xarray\footnote{\texttt{h5py} is a
+  dependency of Xarray} is a widely used Python package for working with
+n-dimensional data. We also tested access times using h5coro, a python
+package optimized for reading HDF5 files from S3 buckets and kerchunk, a
+tool that creates an efficient lookup table for file chunks to allow
+performant partial reads of files.
+
+HDF5 ATL03 files were originally cloud optimized by ``repacking'' them,
+using a relatively new feature in the HDF5 C API called ``paged
+aggregation''. Page aggregation does 2 things: it collects file-level
+metadata from datasets and stores it on dedicated metadata blocks in the
+file; and it forces the library to write data using fixed-size blocks.
+Aggregation allows client libraries to read file metadata with only a
+few requests and uses the page size used in the aggregation as the
+minimal request size, overriding the 1 request per chunk behavior.
+
+\begin{figure*}
+
+\centering{
+
+\includegraphics{figures/figure-2.png}
+
+}
+
+\caption{\label{fig-2}shows how file-level metadata and data gets
+internally packed once we use paged aggregation on a file.}
+
+\end{figure*}%
+
+\section{Results}\label{results}
+
+\begin{figure*}
+
+\centering{
+
+\includegraphics{figures/figure-3.png}
+
+}
+
+\caption{\label{fig-3}Benchmarks show that cloud optimizing ATL03 files
+improved access times at least an order of magnitude when used with
+aligned I/O patterns, this is telling the library about the cloud
+optimization and page size.}
+
+\end{figure*}%
+
+\section{Recommendations}\label{recommendations}
+
+We have split our recommendations for the ATL03 product into 3 main
+categories, creating the files, accessing the files, and future tool
+development.
+
+\subsection{Recommended cloud
+optimizations}\label{recommended-cloud-optimizations}
+
+Based on our testing we recommend the following cloud optimizations for
+creating HDF5 files for the ATL03 product: Create HDF5 files using paged
+aggregation by setting HDF5 library parameters:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  File page strategy: H5F\_FSPACE\_STRATEGY\_PAGE
+\item
+  File page size: 8000000 If repacking an existing file, h5repack
+  contains the code to alter tese variables inside the file
+\end{enumerate}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+ \ExtensionTok{h5repack} \AttributeTok{{-}S}\NormalTok{ PAGE }\AttributeTok{{-}G}\NormalTok{ 8000000 input.h5 output.h5}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\setcounter{enumi}{2}
+\tightlist
+\item
+  Avoid using unlimited dimensions when creating variables because the
+  HDF5 API cannot support it inside bffered pages and representation of
+  these variables is not supported by Kerchunk.
+\end{enumerate}
+
+\subsubsection{Reasoning}\label{reasoning}
+
+Based on the variable size of ATL03 it becomes really difficult to
+allocate a fixed metadata page, big files contain north of 30MB of
+metadata, but the median sized file is below 8MB. If we had adopted user
+block we would have caused an increase in the file size and storage cost
+of approximate 30\% (reference to our tests). Another consequence of
+using a dedicated fixed page for file-level metadata is that metadata
+overflow will generate the same impact in access times, the library will
+fetch the metadata in one go but the rest will be using the predefined
+block size of 4kb.
+
+Paged aggregation is thus the simplest way of cloud optimizing an HDF5
+file as the metadata will keep filling dedicated pages until all the
+file-level metadata is stored at the front of the file. Chunk sizes
+cannot be larger than the page size and when chunk sizes are smaller we
+need to take into account how these chunks will fit on a page, in an
+ideal scenario all the space will be filled but that is not the case and
+we will end up with unused space See~\ref{fig-2}.
+
+\subsection{Recommended access
+patterns}\label{recommended-access-patterns}
+
+Placeholder
+
+\subsection{Recommended tooling
+development}\label{recommended-tooling-development}
+
+Placeholder
+
+\subsection{Mission implementation}\label{mission-implementation}
+
+ATL03 is a complex science data product containing both segmented (20
+meters along-track) and large, variable-rate photon datasets. ATL03 is
+created using pipeline-style processing where the science data and
+NetCDF-style metadata are written by independent software packages. The
+following steps were employed to create cloud-optimized Release 007
+ATL03 products, while minimizing increases in file size:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  Set the ``file space strategy'' to H5F\_FSPACE\_STRATEGY\_PAGE and
+  enabled ``free space tracking'' within the HDF5 file creation property
+  list.
+\item
+  Set the ``file space page size'' to 8MiB.
+\item
+  Changed all ``COMPACT'' dataset storage types to ``CONTIGUOUS''.
+\item
+  Increased the ``chunk size'' of the photon-rate datasets (from 10,000
+  to 100,000 elements), while making sure no ``chunk sizes'' exceeded
+  the 8MiB ``file space page size''.
+\item
+  Introduced a new production step that executes the ``h5repack''
+  utility (with no options) to create a ``defragmented'' final product.
+\end{enumerate}
+
+\section{Discussion}\label{discussion}
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  Chunking shapes and sizes
+\item
+  Paged aggregation vs User block
+\item
+  Side effects on different access patterns, e.g.~Kerchunk
+\end{enumerate}
 
@@
 
-\bibitem[\citeproctext]{ref-Hubbard2021}
-Hubbard, B., Christoffersen, P., Doyle, S. H., Chudley, T. R.,
-Schoonman, C. M., Law, R., \& Bougamont, M. (2021). Borehole-based
-characterization of deep mixed-mode crevasses at a greenlandic outlet
-glacier. \emph{{AGU} Advances}, \emph{2}(2).
-\url{https://doi.org/10.1029/2020av000291}
+\bibitem[\citeproctext]{ref-NEUMANN2019111325}
+Neumann, T. A., Martino, A. J., Markus, T., Bae, S., Bock, M. R.,
+Brenner, A. C., et al. (2019). The ice, cloud, and land elevation
+satellite -- 2 mission: A global geolocated photon product derived from
+the advanced topographic laser altimeter system. \emph{Remote Sensing of
+Environment}, \emph{233}, 111325.
+https://doi.org/\url{https://doi.org/10.1016/j.rse.2019.111325}
 

### 2024-09-05 paper.qmd
partial update, expanded methodology

@@
 
-By default, file-level metadata associated with a dataset is stored in chunks of 4kb. This produces a lot of fragmentation across the file especially for data with many variables and nested groups.
+When working with large datasets, especially those that include numerous variables and nested groups, the storage of file-level metadata can become a challenge.
+By default, metadata associated with each dataset is stored in chunks of 4 kilobytes (KB).
+This chunking mechanism was originally intended to optimize storage efficiency and access speed on disks with hardware resources available more than 20 years ago.
+In datasets with many variables and/or complex hierarchical structures, these 4KB chunks can lead to significant fragmentation.
+
+Fragmentation occurs when this metadata is spread out across multiple non-contiguous chunks within the file.
+This results in inefficiencies when accessing or modifying data because compatible libraries need to read from multiple, scattered locations in the file.
+Over time, as the dataset grows and evolves, this fragmentation can compound, leading to degraded performance and increased storage overhead.
+In particular, operations that involve reading or writing metadata, such as opening a file, checking attributes, or modifying variables, can become slower and more resource-intensive.
 
@@
 
-Because of the historical complexity of operations with the HDF5 format, there has been a necessity to make the library thread-safe and similarly to what happens in the Python language, the simplest mechanism to implement this is to have a global API lock.
-This global lock is not as big of an issue when we read data from local disk but it becomes a major bottleneck when we read data over the network because each read is sequential and latency in the cloud is exponentially bigger than local access.
+Because of the historical complexity of operations with the HDF5 format[@The_HDF_Group_Hierarchical_Data_Format], there has been a necessity to make the library thread-safe and similarly to what happens in the Python language, the simplest mechanism to implement this is to have a global API lock.
+This global lock is not as big of an issue when we read data from local disk but it becomes a major bottleneck when we read data over the network because each read is sequential and latency in the cloud is exponentially bigger than local access [@Mozilla-latency-2024] [@scott-2020].
 
@@
 
-#### Background and data selection
+#### **Background and data selection**
 
-As a result of community feedback and “hack weeks” organized by NSIDC and UW eScience Institute in 2023, NSIDC started the Cloud Optimized Format Investigation (COFI) project to improve access to HDF5 from the ICESat-2 mission.
-A spaceborne lidar that retrieves surface topography of the Earth’s ice sheets, land and [oceans @NEUMANN2019111325].
+As a result of community feedback and “hack weeks” organized by NSIDC and UW eScience Institute in 2023[@h5cloud2023], NSIDC started the Cloud Optimized Format Investigation (COFI) project to improve access to HDF5 from the ICESat-2 mission.
+A spaceborne lidar that retrieves surface topography of the Earth’s ice sheets, land and oceans [@NEUMANN2019111325].
 Because of its complexity, large size and importance for cryospheric studies we targeted the ATL03 dataset.
@@
 
-We tested access times to original and cloud-optimized small (1 GB), medium (2 GB) and large (7 GB) HDF5 ATL03 files [list files tested] stored in AWS S3 buckets in region us-west-2, the region hosting NASA’s Earthdata Cloud archives.
-Files were accessed using Python tools commonly used by Earth scientists: h5py and Xarray.
+We tested access times to original and different configurations of cloud-optimized HDF5 ATL03 files [list files tested] stored in AWS S3 buckets in region us-west-2, the region hosting NASA’s Earthdata Cloud archives.
+Files were accessed using Python tools commonly used by Earth scientists: h5py and Xarray[@Hoyer2017-su].
 h5py is a Python wrapper around the HDF5 C API.
@@
 
-HDF5 ATL03 files were originally cloud optimized by  “repacking” them, using a relatively new feature in the HDF5 C API called “paged aggregation”.
-Page aggregation does 2 things: it collects file-level metadata from datasets and stores it on dedicated metadata blocks in the file; and it forces the library to write data using fixed-size blocks.
+The test files were originally cloud optimized by  “repacking” them, using a relatively new feature in the HDF5 C API called “paged aggregation”.
+Page aggregation does 2 things: it collects file-level metadata from datasets and stores it on dedicated metadata blocks in the file; and it forces the library to write both data and metadata using these fixed-size pages.
 Aggregation allows client libraries to read  file metadata with only a few requests and uses the page size used in the aggregation as the minimal request size, overriding the 1 request per chunk behavior.
@@
 
-## Results
+As we can see in [@fig-2], when we cloud optimize a file using paged-aggregation there are some considerations and behavior that we had to take into account. The first thing to observe is that 
+page aggregation will --as we mentioned-- consolidate the file-evel metadata at the front of the file and will add information in the so-called superblock^[The HDF5 superblock is a crucial component of the HDF5 file format, acting as the starting point for accessing all data within the file.
+It stores important metadata such as the version of the file format, pointers to the root group, and addresses for locating different file components]
+The next thing to notice is that page size us uses across the board for metadata and data as of October 2024 and version 1.14 of the HDF5 library, page size cannot dynamically adjust to the total metadata size.
 
@@
 ![](figures/figure-3.png)
+
+shows how file-level metadata and data packing inside aggregated pages leave unused space that can potentially increase the file size in a considerable way.
+
+:::
+
+This one page size for all approach simplifies how the HDF5 API reads the file (if instructucted) but it also brings unused page space and chunk over-reads.
+In the case of the ICESat-2 dataset (ATL03) the data itself has been partitioned and each granule
+represents a segment in the satellite orbit and within the file the most relevant dataset is chunked using 10,000 items per chunk, with data being float-32 and using a fast compression value, the resulting chunk size is on average under 40KB, which is really small
+for an HTTP request, especially when we have to read them sequentially.
+Because of these considerations, we opted for testing different page sizes, and increase chunk size.
+The following table describes the different configurations used in our tests.
+
+| prefix                 	| description                                        	| % file size increase 	| ~km per chunk 	| shape     	| page size 	| avg_chunk_size 	|
+|------------------------	|----------------------------------------------------	|----------------------	|---------------	|-----------	|-----------	|----------------	|
+| original               	| original file from ATL03 v6 (1gb and 7gb)          	| 0                    	| 1.5km         	| (10000,)  	| N/A       	| 35kb           	|
+| original-kerchunk      	| kerchunk sidecar of the original file              	| N/A                  	| 1.5km         	| (10000,)  	| N/A       	| 35kb           	|
+| page-only-4mb          	| paged-aggregated file with 4mb per page            	| ~1%                  	| 1.5km         	| (10000,)  	| 4MB       	| 35kb           	|
+| page-only-8mb          	| paged-aggregated file with 4mb per pag8            	| ~1%                  	| 1.5km         	| (10000,)  	| 8MB       	| 35kb           	|
+| rechunked-4mb          	| page-aggregated and bigger chunk sizes             	| ~1%                  	| 10km          	| (100000,) 	| 4MB       	| 400kb          	|
+| rechunked-8mb          	| page-aggregated and bigger chunk sizes             	| ~1%                  	| 10km          	| (100000,) 	| 8MB       	| 400kb          	|
+| rechunked-8mb-kerchunk 	| kerchunk sidecar of the last paged-aggregated file 	| N/A                  	| 10km          	| (10000,)  	| 8MB       	| 400kb          	|
+
+This table represents the different configurations we used for our tests in 2 file sizes.
+It's worth noticing that we encountered a few outlier cases where compression and chunk sizes led page aggregation to an increase in file size of approximately 10% which was above the desired value for NSIDC (5% max)
+We tested these files using the most common libraries to handle HDF5 and 2 different I/O drivers that support remote access to AWS S3, fsspec and the native S3.
+The results of our testing is explained in the next section and the code
+to reproduce the results is in the attached notebooks.
+
+## Results
+
+::: {#fig-5 fig-env="figure*"}
+
+![](figures/figure-5.png)
 
@@
 
-## References {.unnumbered}
+## References
 

### 2024-09-05 _tex/paper.tex
Built site for gh-pages

@@
 
-By default, file-level metadata associated with a dataset is stored in
-chunks of 4kb. This produces a lot of fragmentation across the file
-especially for data with many variables and nested groups.
+When working with large datasets, especially those that include numerous
+variables and nested groups, the storage of file-level metadata can
+become a challenge. By default, metadata associated with each dataset is
+stored in chunks of 4 kilobytes (KB). This chunking mechanism was
+originally intended to optimize storage efficiency and access speed on
+disks with hardware resources available more than 20 years ago. In
+datasets with many variables and/or complex hierarchical structures,
+these 4KB chunks can lead to significant fragmentation.
+
+Fragmentation occurs when this metadata is spread out across multiple
+non-contiguous chunks within the file. This results in inefficiencies
+when accessing or modifying data because compatible libraries need to
+read from multiple, scattered locations in the file. Over time, as the
+dataset grows and evolves, this fragmentation can compound, leading to
+degraded performance and increased storage overhead. In particular,
+operations that involve reading or writing metadata, such as opening a
+file, checking attributes, or modifying variables, can become slower and
+more resource-intensive.
 
@@
 
-Because of the historical complexity of operations with the HDF5 format,
-there has been a necessity to make the library thread-safe and similarly
-to what happens in the Python language, the simplest mechanism to
-implement this is to have a global API lock. This global lock is not as
-big of an issue when we read data from local disk but it becomes a major
-bottleneck when we read data over the network because each read is
-sequential and latency in the cloud is exponentially bigger than local
-access.
+Because of the historical complexity of operations with the HDF5
+format(The HDF Group, n.d.), there has been a necessity to make the
+library thread-safe and similarly to what happens in the Python
+language, the simplest mechanism to implement this is to have a global
+API lock. This global lock is not as big of an issue when we read data
+from local disk but it becomes a major bottleneck when we read data over
+the network because each read is sequential and latency in the cloud is
+exponentially bigger than local access (MDN, 2024) (Scott, 2020).
 
@@
 
-\subsubsection{Background and data
-selection}\label{background-and-data-selection}
+\subsubsection{\texorpdfstring{\textbf{Background and data
+selection}}{Background and data selection}}\label{background-and-data-selection}
 
 As a result of community feedback and ``hack weeks'' organized by NSIDC
-and UW eScience Institute in 2023, NSIDC started the Cloud Optimized
-Format Investigation (COFI) project to improve access to HDF5 from the
-ICESat-2 mission. A spaceborne lidar that retrieves surface topography
-of the Earth's ice sheets, land and (oceans Neumann et al., 2019).
-Because of its complexity, large size and importance for cryospheric
-studies we targeted the ATL03 dataset. ATL03 core data are geolocated
-photon heights from the ICESat-2 ATLAS instrument. Each file contains
-1003 geophysical variables in 6 data groups. Although our research was
-focused on this dataset, most of our findings are applicable to any
-dataset stored in HDF5 and NetCDF4.
+and UW eScience Institute in 2023(ICESAT-2 HackWeek, 2023), NSIDC
+started the Cloud Optimized Format Investigation (COFI) project to
+improve access to HDF5 from the ICESat-2 mission. A spaceborne lidar
+that retrieves surface topography of the Earth's ice sheets, land and
+oceans (Neumann et al., 2019). Because of its complexity, large size and
+importance for cryospheric studies we targeted the ATL03 dataset. ATL03
+core data are geolocated photon heights from the ICESat-2 ATLAS
+instrument. Each file contains 1003 geophysical variables in 6 data
+groups. Although our research was focused on this dataset, most of our
+findings are applicable to any dataset stored in HDF5 and NetCDF4.
 
@@
 
-We tested access times to original and cloud-optimized small (1 GB),
-medium (2 GB) and large (7 GB) HDF5 ATL03 files {[}list files tested{]}
-stored in AWS S3 buckets in region us-west-2, the region hosting NASA's
-Earthdata Cloud archives. Files were accessed using Python tools
-commonly used by Earth scientists: h5py and Xarray. h5py is a Python
-wrapper around the HDF5 C API. xarray\footnote{\texttt{h5py} is a
+We tested access times to original and different configurations of
+cloud-optimized HDF5 ATL03 files {[}list files tested{]} stored in AWS
+S3 buckets in region us-west-2, the region hosting NASA's Earthdata
+Cloud archives. Files were accessed using Python tools commonly used by
+Earth scientists: h5py and Xarray(Hoyer \& Hamman, 2017). h5py is a
+Python wrapper around the HDF5 C API. xarray\footnote{\texttt{h5py} is a
   dependency of Xarray} is a widely used Python package for working with
@@
 
-HDF5 ATL03 files were originally cloud optimized by ``repacking'' them,
+The test files were originally cloud optimized by ``repacking'' them,
 using a relatively new feature in the HDF5 C API called ``paged
@@
 metadata from datasets and stores it on dedicated metadata blocks in the
-file; and it forces the library to write data using fixed-size blocks.
-Aggregation allows client libraries to read file metadata with only a
-few requests and uses the page size used in the aggregation as the
-minimal request size, overriding the 1 request per chunk behavior.
+file; and it forces the library to write both data and metadata using
+these fixed-size pages. Aggregation allows client libraries to read file
+metadata with only a few requests and uses the page size used in the
+aggregation as the minimal request size, overriding the 1 request per
+chunk behavior.
 
@@
 
-\section{Results}\label{results}
+As we can see in Figure~\ref{fig-2}, when we cloud optimize a file using
+paged-aggregation there are some considerations and behavior that we had
+to take into account. The first thing to observe is that page
+aggregation will --as we mentioned-- consolidate the file-evel metadata
+at the front of the file and will add information in the so-called
+superblock\footnote{The HDF5 superblock is a crucial component of the
+  HDF5 file format, acting as the starting point for accessing all data
+  within the file. It stores important metadata such as the version of
+  the file format, pointers to the root group, and addresses for
+  locating different file components} The next thing to notice is that
+page size us uses across the board for metadata and data as of October
+2024 and version 1.14 of the HDF5 library, page size cannot dynamically
+adjust to the total metadata size.
 
@@
 
-\caption{\label{fig-3}Benchmarks show that cloud optimizing ATL03 files
+\caption{\label{fig-3}shows how file-level metadata and data packing
+inside aggregated pages leave unused space that can potentially increase
+the file size in a considerable way.}
+
+\end{figure*}%
+
+This one page size for all approach simplifies how the HDF5 API reads
+the file (if instructucted) but it also brings unused page space and
+chunk over-reads. In the case of the ICESat-2 dataset (ATL03) the data
+itself has been partitioned and each granule represents a segment in the
+satellite orbit and within the file the most relevant dataset is chunked
+using 10,000 items per chunk, with data being float-32 and using a fast
+compression value, the resulting chunk size is on average under 40KB,
+which is really small for an HTTP request, especially when we have to
+read them sequentially. Because of these considerations, we opted for
+testing different page sizes, and increase chunk size. The following
+table describes the different configurations used in our tests.
+
+\begin{longtable}[]{@{}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1589}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.3444}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1457}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.0993}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.0728}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.0728}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1060}}@{}}
+\toprule\noalign{}
+\begin{minipage}[b]{\linewidth}\raggedright
+prefix
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+description
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+\% file size increase
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+\textasciitilde km per chunk
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+shape
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+page size
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+avg\_chunk\_size
+\end{minipage} \\
+\midrule\noalign{}
+\endhead
+\bottomrule\noalign{}
+\endlastfoot
+original & original file from ATL03 v6 (1gb and 7gb) & 0 & 1.5km &
+(10000,) & N/A & 35kb \\
+original-kerchunk & kerchunk sidecar of the original file & N/A & 1.5km
+& (10000,) & N/A & 35kb \\
+page-only-4mb & paged-aggregated file with 4mb per page &
+\textasciitilde1\% & 1.5km & (10000,) & 4MB & 35kb \\
+page-only-8mb & paged-aggregated file with 4mb per pag8 &
+\textasciitilde1\% & 1.5km & (10000,) & 8MB & 35kb \\
+rechunked-4mb & page-aggregated and bigger chunk sizes &
+\textasciitilde1\% & 10km & (100000,) & 4MB & 400kb \\
+rechunked-8mb & page-aggregated and bigger chunk sizes &
+\textasciitilde1\% & 10km & (100000,) & 8MB & 400kb \\
+rechunked-8mb-kerchunk & kerchunk sidecar of the last paged-aggregated
+file & N/A & 10km & (10000,) & 8MB & 400kb \\
+\end{longtable}
+
+This table represents the different configurations we used for our tests
+in 2 file sizes. It's worth noticing that we encountered a few outlier
+cases where compression and chunk sizes led page aggregation to an
+increase in file size of approximately 10\% which was above the desired
+value for NSIDC (5\% max) We tested these files using the most common
+libraries to handle HDF5 and 2 different I/O drivers that support remote
+access to AWS S3, fsspec and the native S3. The results of our testing
+is explained in the next section and the code to reproduce the results
+is in the attached notebooks.
+
+\section{Results}\label{results}
+
+\begin{figure*}
+
+\centering{
+
+\includegraphics{figures/figure-5.png}
+
+}
+
+\caption{\label{fig-5}Benchmarks show that cloud optimizing ATL03 files
 improved access times at least an order of magnitude when used with
@@
 
-\section*{References}\label{references}
-\addcontentsline{toc}{section}{References}
+\section{References}\label{references}
 
@@
 \vspace{1em}
+
+\bibitem[\citeproctext]{ref-Hoyer2017-su}
+Hoyer, S., \& Hamman, J. (2017). Xarray: {N-D} labeled arrays and
+datasets in python. \emph{J. Open Res. Softw.}, \emph{5}(1), 10.
+
+\bibitem[\citeproctext]{ref-h5cloud2023}
+ICESAT-2 HackWeek, H. C. (2023). h5cloud: Tools for cloud-based analysis
+of HDF5 data (Version v1.0.0). Retrieved from
+\url{https://github.com/ICESAT-2HackWeek/h5cloud}
+
+\bibitem[\citeproctext]{ref-Mozilla-latency-2024}
+MDN, M. (2024, May). Understanding latency. Retrieved from
+\url{https://developer.mozilla.org/en-US/docs/Web/Performance/Understanding_latency}
 
@@
 
+\bibitem[\citeproctext]{ref-scott-2020}
+Scott, C. (2020). Numbers every programmer should know. Retrieved from
+\url{https://colin-scott.github.io/personal_website/research/interactive_latency.html}
+
+\bibitem[\citeproctext]{ref-The_HDF_Group_Hierarchical_Data_Format}
+The HDF Group. (n.d.). {Hierarchical Data Format, version 5}. Retrieved
+from \url{https://github.com/HDFGroup/hdf5}
+
 \end{CSLReferences}

### 2024-10-03 paper.qmd
another incremental update, maybe we can finish this next week

@@
 
-As a result of community feedback and “hack weeks” organized by NSIDC and UW eScience Institute in 2023[@h5cloud2023], NSIDC started the Cloud Optimized Format Investigation (COFI) project to improve access to HDF5 from the ICESat-2 mission.
-A spaceborne lidar that retrieves surface topography of the Earth’s ice sheets, land and oceans [@NEUMANN2019111325].
-Because of its complexity, large size and importance for cryospheric studies we targeted the ATL03 dataset.
-ATL03 core data are geolocated photon heights from the ICESat-2 ATLAS instrument.
-Each file contains 1003 geophysical variables in 6 data groups.
+As a result of community feedback and “hack weeks” organized by NSIDC and UW eScience Institute in 2023[@h5cloud2023], NSIDC started the Cloud Optimized Format Investigation (COFI) project to improve access to HDF5 from the ICESat-2 mission, a spaceborne lidar that retrieves surface topography of the Earth’s ice sheets, land and oceans [@NEUMANN2019111325].
+Because of its complexity, large size and importance for cryospheric studies we targeted the ATL03 data product.
+The most relevant variable in ATL03 are geolocated photon heights from the ICESat-2 ATLAS instrument.
+Each ATL03 file contains 1003 geophysical variables in 6 data groups.
 Although our research was focused on this dataset, most of our findings are applicable to any dataset stored in HDF5 and NetCDF4.
@@
 
-We tested access times to original and different configurations of cloud-optimized HDF5 ATL03 files [list files tested] stored in AWS S3 buckets in region us-west-2, the region hosting NASA’s Earthdata Cloud archives.
+We tested access times to original and different configurations of cloud-optimized HDF5 [ATL03 files](https://its-live-data.s3.amazonaws.com/index.html#test-space/cloud-experiments/h5cloud/) stored in AWS S3 buckets in region us-west-2, the region hosting NASA’s Earthdata Cloud archives.
 Files were accessed using Python tools commonly used by Earth scientists: h5py and Xarray[@Hoyer2017-su].
@@
 The test files were originally cloud optimized by  “repacking” them, using a relatively new feature in the HDF5 C API called “paged aggregation”.
-Page aggregation does 2 things: it collects file-level metadata from datasets and stores it on dedicated metadata blocks in the file; and it forces the library to write both data and metadata using these fixed-size pages.
-Aggregation allows client libraries to read  file metadata with only a few requests and uses the page size used in the aggregation as the minimal request size, overriding the 1 request per chunk behavior.
+Page aggregation does 2 things: fisrt, it collects file-level metadata from datasets and stores it on dedicated metadata blocks at the front of the file; second, it forces the library to write both data and metadata using these fixed-size pages.
+Aggregation allows client libraries to read  file metadata with only a few requests using the page size as a fixed request size, overriding the 1 request per chunk behavior.
 
@@
 | rechunked-8mb          	| page-aggregated and bigger chunk sizes             	| ~1%                  	| 10km          	| (100000,) 	| 8MB       	| 400kb          	|
-| rechunked-8mb-kerchunk 	| kerchunk sidecar of the last paged-aggregated file 	| N/A                  	| 10km          	| (10000,)  	| 8MB       	| 400kb          	|
+| rechunked-8mb-kerchunk 	| kerchunk sidecar of the last paged-aggregated file 	| N/A                  	| 10km          	| (100000,)  	| 8MB       	| 400kb          	|
 
@@
 
+::: {#fig-4 fig-env="figure*"}
+
+![](figures/figure-4.png)
+
+shows that using paged aggregation alone is not a complete solution. This behavior us caused by over-reads of data now distributed in pages and the internals of HDF5 not knowing how to optimize 
+the requests.
+This means that if we cloud optimize alone and use the same code, in some cases we'll make access to these files even slower.
+A very important thing to notice here is that rechunking the file, in this case using 10X bigger chunks results in a predictable 10X improvement in access times without any cloud optimization involved.
+Having less chunks generates less metadata and bigger requests, in general is it recommended that chunk sizes should range between 1MB and 10MB[Add citation, S3 and HDF5] and if we have anough memory and bandwith even
+bigger (Pangeo recommends up to 100MB chunks)[Add citation.]
+
+:::
+
 ::: {#fig-5 fig-env="figure*"}
@@
 
-Benchmarks show that cloud optimizing ATL03 files improved access times at least an order of magnitude when used with aligned I/O patterns, this is telling the library about the cloud optimization and page size.
+shows that performance once the I/O configuration is aligned with the chunking in the file, access times perform on par with cloud optimized access patterns like Kerchunk/Zarr. 
+These numbers are from in-region execution. Out of region is considerable slower for the non cloud optimized case.
 
@@
 
-Placeholder
+In progress
 

### 2024-10-03 notebooks/h5logger.py

@@ -1,0 +1,140 @@
+import re
+import numpy as np
+import pandas as pd
+import logging
+import s3fs
+import fsspec
+import time
+import h5py
+from datetime import datetime
+from uuid import uuid4
+
+
+def parse_fsspec_log(log_path):
+    """
+    This method only parses fsspec logs that have a FileSize: attached to them.
+    """
+    head_line  = re.compile('<File-like object S3FileSystem, .*?>\s*(read: 0 - \d+)')
+    fsize_line = re.compile('FileSize: (\d+)')
+    # range_line = re.compile('<File-like object S3FileSystem, .*?>\s* read: (?P<start>[0-9]+) - (?P<end>[0-9]+)')
+    range_line = re.compile('<File-like object S3FileSystem, .*?>\s* read: (?P<start>[0-9]+) - (?P<end>[0-9]+)(?:\s*,\s*.*?:\s*(?P<hits>[0-9]+)\s*hits,\s*(?P<misses>[0-9]+)\s*misses)?')
+  
+
+
+    ranges = list()
+    with open(log_path) as logtxt:
+        for line in logtxt:
+            if head_line.match(line):
+                break
+        else:
+            raise RuntimeError('HEAD line not found in the log file')
+
+        for line in logtxt:
+            match = fsize_line.match(line)
+            if match:
+                fsize = int(match.group(1))
+                break
+        else:
+            raise RuntimeError('FILESIZE line not found in the log file')
+            
+        logtxt.seek(0)
+        for line in logtxt:
+            match = range_line.match(line)
+            if match:
+                start=int(match.group(1))
+                end=int(match.group(2))
+                hits=match.group(3)
+                missed=match.group(4)
+                rsize=end-start+1
+                
+                ranges.append({"start": start, "end": end, "size": rsize, "hits": hits, "missed": missed})
+
+    df = pd.DataFrame(ranges, columns=['start', 'end', 'size', 'hits', 'missed'])
+    return df
+
+def read_file(info):
+    h5py_fsspec_benchmarks = {}
+    ranges = None
+    file_size = None
+    block_size = None
+    iteration, dataset, variables, flavor, url, optimized_read, driver, default_io_params, optimized_io_params = info
+    if url.endswith(".json"):
+        return {}   
+    io_params = default_io_params
+    if optimized_read:
+        if "rechunked" in url or "page" in url:
+            optimized = "yes"
+            print(f"Reading: {url} with optimized I/O parameters")
[…]

### 2024-10-03 notebooks/optimize-atl06.py

@@ -1,0 +1,573 @@
+#!/usr/bin/env pythonw
+"""Creates a cloud-optimized version of an ICESat-2 HDF5 file."""
+
+# Copyright (c) 2024, University of Washington
+# All rights reserved.
+#
+# Redistribution and use in source and binary forms, with or without
+# modification, are permitted provided that the following conditions are met:
+#
+# 1. Redistributions of source code must retain the above copyright notice,
+#    this list of conditions and the following disclaimer.
+#
+# 2. Redistributions in binary form must reproduce the above copyright notice,
+#    this list of conditions and the following disclaimer in the documentation
+#    and/or other materials provided with the distribution.
+#
+# 3. Neither the name of the University of Washington nor the names of its
+#    contributors may be used to endorse or promote products derived from this
+#    software without specific prior written permission.
+#
+# THIS SOFTWARE IS PROVIDED BY THE UNIVERSITY OF WASHINGTON AND CONTRIBUTORS
+# “AS IS” AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT LIMITED
+# TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR A PARTICULAR
+# PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE UNIVERSITY OF WASHINGTON OR
+# CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL, SPECIAL,
+# EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT LIMITED TO,
+# PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE, DATA, OR PROFITS;
+# OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY THEORY OF LIABILITY,
+# WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT (INCLUDING NEGLIGENCE OR
+# OTHERWISE) ARISING IN ANY WAY OUT OF THE USE OF THIS SOFTWARE, EVEN IF
+# ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
+
+import os
+import sys
+import argparse
+import traceback
+import h5py
+from h5py import h5p, h5f
+import numpy as np
+import time
+
+__version_ = "1.0.1"
+
+
[…]

### 2024-10-03 _tex/paper.tex
Built site for gh-pages

@@
 started the Cloud Optimized Format Investigation (COFI) project to
-improve access to HDF5 from the ICESat-2 mission. A spaceborne lidar
+improve access to HDF5 from the ICESat-2 mission, a spaceborne lidar
 that retrieves surface topography of the Earth's ice sheets, land and
 oceans (Neumann et al., 2019). Because of its complexity, large size and
-importance for cryospheric studies we targeted the ATL03 dataset. ATL03
-core data are geolocated photon heights from the ICESat-2 ATLAS
-instrument. Each file contains 1003 geophysical variables in 6 data
-groups. Although our research was focused on this dataset, most of our
-findings are applicable to any dataset stored in HDF5 and NetCDF4.
+importance for cryospheric studies we targeted the ATL03 data product.
+The most relevant variable in ATL03 are geolocated photon heights from
+the ICESat-2 ATLAS instrument. Each ATL03 file contains 1003 geophysical
+variables in 6 data groups. Although our research was focused on this
+dataset, most of our findings are applicable to any dataset stored in
+HDF5 and NetCDF4.
 
@@
 We tested access times to original and different configurations of
-cloud-optimized HDF5 ATL03 files {[}list files tested{]} stored in AWS
-S3 buckets in region us-west-2, the region hosting NASA's Earthdata
-Cloud archives. Files were accessed using Python tools commonly used by
-Earth scientists: h5py and Xarray(Hoyer \& Hamman, 2017). h5py is a
-Python wrapper around the HDF5 C API. xarray\footnote{\texttt{h5py} is a
-  dependency of Xarray} is a widely used Python package for working with
-n-dimensional data. We also tested access times using h5coro, a python
-package optimized for reading HDF5 files from S3 buckets and kerchunk, a
-tool that creates an efficient lookup table for file chunks to allow
-performant partial reads of files.
+cloud-optimized HDF5
+\href{https://its-live-data.s3.amazonaws.com/index.html\#test-space/cloud-experiments/h5cloud/}{ATL03
+files} stored in AWS S3 buckets in region us-west-2, the region hosting
+NASA's Earthdata Cloud archives. Files were accessed using Python tools
+commonly used by Earth scientists: h5py and Xarray(Hoyer \& Hamman,
+2017). h5py is a Python wrapper around the HDF5 C API. xarray\footnote{\texttt{h5py}
+  is a dependency of Xarray} is a widely used Python package for working
+with n-dimensional data. We also tested access times using h5coro, a
+python package optimized for reading HDF5 files from S3 buckets and
+kerchunk, a tool that creates an efficient lookup table for file chunks
+to allow performant partial reads of files.
 
@@
 using a relatively new feature in the HDF5 C API called ``paged
-aggregation''. Page aggregation does 2 things: it collects file-level
-metadata from datasets and stores it on dedicated metadata blocks in the
-file; and it forces the library to write both data and metadata using
-these fixed-size pages. Aggregation allows client libraries to read file
-metadata with only a few requests and uses the page size used in the
-aggregation as the minimal request size, overriding the 1 request per
+aggregation''. Page aggregation does 2 things: fisrt, it collects
+file-level metadata from datasets and stores it on dedicated metadata
+blocks at the front of the file; second, it forces the library to write
+both data and metadata using these fixed-size pages. Aggregation allows
+client libraries to read file metadata with only a few requests using
+the page size as a fixed request size, overriding the 1 request per
 chunk behavior.
@@
 rechunked-8mb-kerchunk & kerchunk sidecar of the last paged-aggregated
-file & N/A & 10km & (10000,) & 8MB & 400kb \\
+file & N/A & 10km & (100000,) & 8MB & 400kb \\
 \end{longtable}
@@
 
+\includegraphics{figures/figure-4.png}
+
+}
+
+\caption{\label{fig-4}shows that using paged aggregation alone is not a
+complete solution. This behavior us caused by over-reads of data now
+distributed in pages and the internals of HDF5 not knowing how to
+optimize the requests. This means that if we cloud optimize alone and
+use the same code, in some cases we'll make access to these files even
+slower. A very important thing to notice here is that rechunking the
+file, in this case using 10X bigger chunks results in a predictable 10X
+improvement in access times without any cloud optimization involved.
+Having less chunks generates less metadata and bigger requests, in
+general is it recommended that chunk sizes should range between 1MB and
+10MB{[}Add citation, S3 and HDF5{]} and if we have anough memory and
+bandwith even bigger (Pangeo recommends up to 100MB chunks){[}Add
+citation.{]}}
+
+\end{figure*}%
+
+\begin{figure*}
+
+\centering{
+
 \includegraphics{figures/figure-5.png}
@@
 
-\caption{\label{fig-5}Benchmarks show that cloud optimizing ATL03 files
-improved access times at least an order of magnitude when used with
-aligned I/O patterns, this is telling the library about the cloud
-optimization and page size.}
+\caption{\label{fig-5}shows that performance once the I/O configuration
+is aligned with the chunking in the file, access times perform on par
+with cloud optimized access patterns like Kerchunk/Zarr. These numbers
+are from in-region execution. Out of region is considerable slower for
+the non cloud optimized case.}
 

### 2024-10-09 paper.qmd
fixing typos

@@
 The most recent version of the Hierarchical Data Format is HDF5, a common archival format for n-dimensional scientific data; it has been utilized to store valuable information from astrophysics to earth sciences and everything in between.
-As flexible and powerful as HDF5 can be, it comes with big tradeoffs when it’s accessed from remote storage systems.
+As flexible and powerful as HDF5 can be, it comes with big trade-offs when it’s accessed from remote storage systems.
 
@@
 The test files were originally cloud optimized by  “repacking” them, using a relatively new feature in the HDF5 C API called “paged aggregation”.
-Page aggregation does 2 things: fisrt, it collects file-level metadata from datasets and stores it on dedicated metadata blocks at the front of the file; second, it forces the library to write both data and metadata using these fixed-size pages.
+Page aggregation does 2 things: first, it collects file-level metadata from datasets and stores it on dedicated metadata blocks at the front of the file; second, it forces the library to write both data and metadata using these fixed-size pages.
 Aggregation allows client libraries to read  file metadata with only a few requests using the page size as a fixed request size, overriding the 1 request per chunk behavior.
@@
 As we can see in [@fig-2], when we cloud optimize a file using paged-aggregation there are some considerations and behavior that we had to take into account. The first thing to observe is that 
-page aggregation will --as we mentioned-- consolidate the file-evel metadata at the front of the file and will add information in the so-called superblock^[The HDF5 superblock is a crucial component of the HDF5 file format, acting as the starting point for accessing all data within the file.
+page aggregation will --as we mentioned-- consolidate the file-level metadata at the front of the file and will add information in the so-called superblock^[The HDF5 superblock is a crucial component of the HDF5 file format, acting as the starting point for accessing all data within the file.
 It stores important metadata such as the version of the file format, pointers to the root group, and addresses for locating different file components]
@@
 
-This one page size for all approach simplifies how the HDF5 API reads the file (if instructucted) but it also brings unused page space and chunk over-reads.
+This one page size for all approach simplifies how the HDF5 API reads the file (if configured) but it also brings unused page space and chunk over-reads.
 In the case of the ICESat-2 dataset (ATL03) the data itself has been partitioned and each granule
@@
 A very important thing to notice here is that rechunking the file, in this case using 10X bigger chunks results in a predictable 10X improvement in access times without any cloud optimization involved.
-Having less chunks generates less metadata and bigger requests, in general is it recommended that chunk sizes should range between 1MB and 10MB[Add citation, S3 and HDF5] and if we have anough memory and bandwith even
+Having less chunks generates less metadata and bigger requests, in general is it recommended that chunk sizes should range between 1MB and 10MB[Add citation, S3 and HDF5] and if we have enough memory and bandwidth even
 bigger (Pangeo recommends up to 100MB chunks)[Add citation.]
@@
 2. File page size: 8000000
-  If repacking an existing file, h5repack contains the code to alter tese variables inside the file 
+  If repacking an existing file, h5repack contains the code to alter these variables inside the file 
   ```bash
@@
   ```
-3. Avoid using unlimited dimensions when creating variables because the HDF5 API cannot support it inside bffered pages and representation of these variables is not supported by Kerchunk.
+3. Avoid using unlimited dimensions when creating variables because the HDF5 API cannot support it inside buffered pages and representation of these variables is not supported by Kerchunk.
 

### 2024-10-09 _tex/paper.tex
Built site for gh-pages

@@
 sciences and everything in between. As flexible and powerful as HDF5 can
-be, it comes with big tradeoffs when it's accessed from remote storage
+be, it comes with big trade-offs when it's accessed from remote storage
 systems.
@@
 using a relatively new feature in the HDF5 C API called ``paged
-aggregation''. Page aggregation does 2 things: fisrt, it collects
+aggregation''. Page aggregation does 2 things: first, it collects
 file-level metadata from datasets and stores it on dedicated metadata
@@
 to take into account. The first thing to observe is that page
-aggregation will --as we mentioned-- consolidate the file-evel metadata
+aggregation will --as we mentioned-- consolidate the file-level metadata
 at the front of the file and will add information in the so-called
@@
 This one page size for all approach simplifies how the HDF5 API reads
-the file (if instructucted) but it also brings unused page space and
-chunk over-reads. In the case of the ICESat-2 dataset (ATL03) the data
-itself has been partitioned and each granule represents a segment in the
+the file (if configured) but it also brings unused page space and chunk
+over-reads. In the case of the ICESat-2 dataset (ATL03) the data itself
+has been partitioned and each granule represents a segment in the
 satellite orbit and within the file the most relevant dataset is chunked
@@
 general is it recommended that chunk sizes should range between 1MB and
-10MB{[}Add citation, S3 and HDF5{]} and if we have anough memory and
-bandwith even bigger (Pangeo recommends up to 100MB chunks){[}Add
+10MB{[}Add citation, S3 and HDF5{]} and if we have enough memory and
+bandwidth even bigger (Pangeo recommends up to 100MB chunks){[}Add
 citation.{]}}
@@
   File page size: 8000000 If repacking an existing file, h5repack
-  contains the code to alter tese variables inside the file
+  contains the code to alter these variables inside the file
 \end{enumerate}
@@
   Avoid using unlimited dimensions when creating variables because the
-  HDF5 API cannot support it inside bffered pages and representation of
+  HDF5 API cannot support it inside buffered pages and representation of
   these variables is not supported by Kerchunk.

### 2024-12-10 paper.qmd
addind content

@@
 
-We have split our recommendations for the ATL03 product into 3 main categories, creating the files, accessing the files, and future tool development.
+Based on the benckmarks we got from our tests, we have split our recommendations for the ATL03 product into 3 main categories, creating the files, accessing the files, and future tool development.
+These recommendations aim to streamline HDF5 workflows in cloud environments, enhancing performance and reducing costs.
 
@@
 
-### Recommended access patterns
+### Recommended Access Patterns
 
-In progress
+As we saw in our benchmarks, efficient access to cloud optimized HDF5 files in cloud storage requires that we also optimize our access patterns.
+The following recommendations focus on optimizing workflows for Python users.
+However, these recommendations should be applicable across programming languages.
+It's also worth mentioning that the HDF Group aims to include some of these features in their roadmap.
 
-### Recommended tooling development
+- **Efficient Reads**: Efficiently reading cloud-hosted HDF5 files involves minimizing network requests and prioritizing large sequential reads.
+Configure chunk sizes between 1–10 MB to match the block sizes used in cloud object storage systems, ensuring meaningful data retrieval in each read.
+Avoid small chunks, as they cause excessive HTTP overhead and slower access speeds.
+- **Parallel Access**: Use parallel computing frameworks like `Dask` or multiprocessing to divide read operations across multiple processes or nodes.
+This alleviates the sequential access bottleneck caused by the HDF5 global lock, particularly in workflows accessing multiple datasets.
+- **Cache Management**: Implement caching for metadata to avoid repetitive fetches.
+Tools like `fsspec` or `h5coro` allow in-memory or on-disk caching for frequently accessed data, reducing latency during high-frequency
+- **Regional Access**: Operate workflows in the same cloud region as the data to minimize costs and latency.
+Cross-region data transfer is expensive and introduces significant delays.
+Where possible, deploy virtual machines close to the data storage region.
 
-In progress
+### Recommended Tooling Development
+
+To enable widespread and efficient use of HDF5 files in cloud environments, it is crucial to develop robust tools across all major programming languages.
+The HDF Group has expressed intentions to include these features in their roadmap, ensuring seamless compatibility with emerging cloud storage and computing standards.
+This section highlights tooling strategies to support metadata indexing, driver enhancements, and diagnostics, applicable to Python and other languages.
+
+- **Enhanced HDF5 Drivers:** Improve drivers like `h5py` and `ROS3` to better handle cloud object storage's nuances, such as intelligent request batching and speculative reads.
+This mitigates inefficiencies caused by high-latency networks.
+- **Metadata Indexing:** Develop tools for pre-indexing metadata, similar to Kerchunk.
+These tools should enable clients to retrieve only necessary data offsets, avoiding full metadata reads and improving access times.
+- **Kerchunk-like Integration:** Extend Kerchunk to integrate seamlessly with analysis libraries like Xarray.
+This includes building robust sidecar files that efficiently map hierarchical datasets, enabling faster partial reads and enhancing cloud-native workflows.
+- **Diagnostic Tools:** Create tools for diagnostics and performance profiling tailored to cloud-optimized HDF5 files.
+These tools should identify bottlenecks in access patterns and recommend adjustments in configurations or chunking strategies.
 
@@
 
-## Discussion
+### Discussion and Further Work
 
-1. Chunking shapes and sizes
-2. Paged aggregation vs User block
-3. Side effects on different access patterns, e.g. Kerchunk
+We believe that implementing cloud optimized HDF5 will greatly improve downstream workflows that will unlock science in the cloud.
+We also recognize that in order to get there, some key factors in the ecosystem need to be addressed.
+Chunking strategies, adaptive caching and automatic driver configurations should be developed to optimize performance.
+
+Efforts should expand multi-language support, creating universal interfaces and libraries for broader adoption beyond Python.
+Cloud-native enhancements must focus on optimizing HDF5 for distributed systems and object storage, addressing egress costs, ease of use and scalability.
+Finally, advancing ecosystem interoperability involves setting integration standards and aligning with emerging trends such as serverless and edge computing.
+These efforts, combined with community collaboration, will modernize HDF5 to meet the challenges of evolving data-intensive applications.
+
+#### Chunking Shapes and Sizes
+
+Optimizing chunk shapes and sizes is essential for efficient HDF5 usage, especially in cloud environments:
+
+- **Chunk Shape:** Align chunk dimensions with anticipated access patterns. For example, row-oriented queries benefit from row-aligned chunks. 
+- **Chunk Size:** Use chunk sizes between 1–10 MB to match cloud storage block sizes.
+Larger chunks improve sequential access but require more memory.
+Smaller chunks support granular reads but may increase network overhead.
+
+Finally, we recognize that this study has not been as extensive as it could have been (cross language, multiple datasets) and yet we think we ran into the key scenarios data producers will face when they start producing cloud optimized HDf5 files.
+We think that there is room for improvement and experimentation with various configurations based on real-world scenarios is crucial to determine the best performance.
 

### 2024-12-10 _tex/paper.tex
Built site for gh-pages

@@
 \makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+\newsavebox\pandoc@box
+\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
+  \sbox\pandoc@box{#1}%
+  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
+  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
+  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
+  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
+  \else\usebox{\pandoc@box}%
+  \fi%
+}
 % Set default figure placement to htbp
-\makeatletter
 \def\fps@figure{htbp}
@@
 
-\ifLuaTeX
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
 \usepackage{bookmark}
@@
 
-\includegraphics{figures/figure-1.png}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/figure-1.png}}
 
@@
 
-\includegraphics{figures/figure-2.png}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/figure-2.png}}
 
@@
 
-\includegraphics{figures/figure-3.png}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/figure-3.png}}
 
@@
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1589}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.3444}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1457}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.0993}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.0728}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.0728}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1060}}@{}}
+  >{\raggedright\arraybackslash}p{(\linewidth - 12\tabcolsep) * \real{0.1589}}
+  >{\raggedright\arraybackslash}p{(\linewidth - 12\tabcolsep) * \real{0.3444}}
+  >{\raggedright\arraybackslash}p{(\linewidth - 12\tabcolsep) * \real{0.1457}}
+  >{\raggedright\arraybackslash}p{(\linewidth - 12\tabcolsep) * \real{0.0993}}
+  >{\raggedright\arraybackslash}p{(\linewidth - 12\tabcolsep) * \real{0.0728}}
+  >{\raggedright\arraybackslash}p{(\linewidth - 12\tabcolsep) * \real{0.0728}}
+  >{\raggedright\arraybackslash}p{(\linewidth - 12\tabcolsep) * \real{0.1060}}@{}}
 \toprule\noalign{}
@@
 
-\includegraphics{figures/figure-4.png}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/figure-4.png}}
 
@@
 
-\includegraphics{figures/figure-5.png}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/figure-5.png}}
 
@@
 
-We have split our recommendations for the ATL03 product into 3 main
-categories, creating the files, accessing the files, and future tool
-development.
+Based on the benckmarks we got from our tests, we have split our
+recommendations for the ATL03 product into 3 main categories, creating
+the files, accessing the files, and future tool development. These
+recommendations aim to streamline HDF5 workflows in cloud environments,
+enhancing performance and reducing costs.
 
@@
 
-\subsection{Recommended access
-patterns}\label{recommended-access-patterns}
+\subsection{Recommended Access
+Patterns}\label{recommended-access-patterns}
 
-In progress
+As we saw in our benchmarks, efficient access to cloud optimized HDF5
+files in cloud storage requires that we also optimize our access
+patterns. The following recommendations focus on optimizing workflows
+for Python users. However, these recommendations should be applicable
+across programming languages. It's also worth mentioning that the HDF
+Group aims to include some of these features in their roadmap.
 
-\subsection{Recommended tooling
-development}\label{recommended-tooling-development}
+\begin{itemize}
+\tightlist
+\item
+  \textbf{Efficient Reads}: Efficiently reading cloud-hosted HDF5 files
+  involves minimizing network requests and prioritizing large sequential
+  reads. Configure chunk sizes between 1--10 MB to match the block sizes
+  used in cloud object storage systems, ensuring meaningful data
+  retrieval in each read. Avoid small chunks, as they cause excessive
+  HTTP overhead and slower access speeds.
+\item
+  \textbf{Parallel Access}: Use parallel computing frameworks like
+  \texttt{Dask} or multiprocessing to divide read operations across
+  multiple processes or nodes. This alleviates the sequential access
+  bottleneck caused by the HDF5 global lock, particularly in workflows
+  accessing multiple datasets.
+\item
+  \textbf{Cache Management}: Implement caching for metadata to avoid
+  repetitive fetches. Tools like \texttt{fsspec} or \texttt{h5coro}
+  allow in-memory or on-disk caching for frequently accessed data,
+  reducing latency during high-frequency
+\item
+  \textbf{Regional Access}: Operate workflows in the same cloud region
+  as the data to minimize costs and latency. Cross-region data transfer
+  is expensive and introduces significant delays. Where possible, deploy
+  virtual machines close to the data storage region.
+\end{itemize}
 
-In progress
+\subsection{Recommended Tooling
+Development}\label{recommended-tooling-development}
+
+To enable widespread and efficient use of HDF5 files in cloud
+environments, it is crucial to develop robust tools across all major
+programming languages. The HDF Group has expressed intentions to include
+these features in their roadmap, ensuring seamless compatibility with
+emerging cloud storage and computing standards. This section highlights
+tooling strategies to support metadata indexing, driver enhancements,
+and diagnostics, applicable to Python and other languages.
+
+\begin{itemize}
+\tightlist
+\item
+  \textbf{Enhanced HDF5 Drivers:} Improve drivers like \texttt{h5py} and
+  \texttt{ROS3} to better handle cloud object storage's nuances, such as
+  intelligent request batching and speculative reads. This mitigates
+  inefficiencies caused by high-latency networks.
+\item
+  \textbf{Metadata Indexing:} Develop tools for pre-indexing metadata,
+  similar to Kerchunk. These tools should enable clients to retrieve
+  only necessary data offsets, avoiding full metadata reads and
+  improving access times.
+\item
+  \textbf{Kerchunk-like Integration:} Extend Kerchunk to integrate
+  seamlessly with analysis libraries like Xarray. This includes building
+  robust sidecar files that efficiently map hierarchical datasets,
+  enabling faster partial reads and enhancing cloud-native workflows.
+\item
+  \textbf{Diagnostic Tools:} Create tools for diagnostics and
+  performance profiling tailored to cloud-optimized HDF5 files. These
+  tools should identify bottlenecks in access patterns and recommend
+  adjustments in configurations or chunking strategies.
+\end{itemize}
 
@@
 
-\section{Discussion}\label{discussion}
+\subsection{Discussion and Further
+Work}\label{discussion-and-further-work}
 
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
+We believe that implementing cloud optimized HDF5 will greatly improve
+downstream workflows that will unlock science in the cloud. We also
+recognize that in order to get there, some key factors in the ecosystem
+need to be addressed. Chunking strategies, adaptive caching and
+automatic driver configurations should be developed to optimize
+performance.
+
+Efforts should expand multi-language support, creating universal
+interfaces and libraries for broader adoption beyond Python.
+Cloud-native enhancements must focus on optimizing HDF5 for distributed
+systems and object storage, addressing egress costs, ease of use and
+scalability. Finally, advancing ecosystem interoperability involves
+setting integration standards and aligning with emerging trends such as
+serverless and edge computing. These efforts, combined with community
+collaboration, will modernize HDF5 to meet the challenges of evolving
+data-intensive applications.
+
+\subsubsection{Chunking Shapes and
+Sizes}\label{chunking-shapes-and-sizes}
+
+Optimizing chunk shapes and sizes is essential for efficient HDF5 usage,
+especially in cloud environments:
+
+\begin{itemize}
 \tightlist
 \item
-  Chunking shapes and sizes
+  \textbf{Chunk Shape:} Align chunk dimensions with anticipated access
+  patterns. For example, row-oriented queries benefit from row-aligned
+  chunks.
 \item
-  Paged aggregation vs User block
-\item
-  Side effects on different access patterns, e.g.~Kerchunk
-\end{enumerate}
+  \textbf{Chunk Size:} Use chunk sizes between 1--10 MB to match cloud
+  storage block sizes. Larger chunks improve sequential access but
+  require more memory. Smaller chunks support granular reads but may
+  increase network overhead.
+\end{itemize}
+
+Finally, we recognize that this study has not been as extensive as it
+could have been (cross language, multiple datasets) and yet we think we
+ran into the key scenarios data producers will face when they start
+producing cloud optimized HDf5 files. We think that there is room for
+improvement and experimentation with various configurations based on
+real-world scenarios is crucial to determine the best performance.
 

### 2024-12-18 paper.qmd
Update paper.qmd

@@
 As we can see in [@fig-2], when we cloud optimize a file using paged-aggregation there are some considerations and behavior that we had to take into account. The first thing to observe is that 
-page aggregation will --as we mentioned-- consolidate the file-level metadata at the front of the file and will add information in the so-called superblock^[The HDF5 superblock is a crucial component of the HDF5 file format, acting as the starting point for accessing all data within the file.
+page aggregation will -- as we mentioned -- consolidate the file-level metadata at the front of the file and will add information in the so-called superblock^[The HDF5 superblock is a crucial component of the HDF5 file format, acting as the starting point for accessing all data within the file.
 It stores important metadata such as the version of the file format, pointers to the root group, and addresses for locating different file components]
-The next thing to notice is that page size us uses across the board for metadata and data as of October 2024 and version 1.14 of the HDF5 library, page size cannot dynamically adjust to the total metadata size.
+The next thing to notice is page size is used across the board for metadata and data as of October 2024 and version 1.14 of the HDF5 library, page size cannot dynamically adjust to the total metadata size.
 
@@
 This table represents the different configurations we used for our tests in 2 file sizes.
-It's worth noticing that we encountered a few outlier cases where compression and chunk sizes led page aggregation to an increase in file size of approximately 10% which was above the desired value for NSIDC (5% max)
+It's worth noticing we encountered a few outlier cases where compression and chunk sizes led page aggregation to an increase in file size of approximately 10% which was above the desired value for NSIDC (5% max)
 We tested these files using the most common libraries to handle HDF5 and 2 different I/O drivers that support remote access to AWS S3, fsspec and the native S3.
-The results of our testing is explained in the next section and the code
+The results of our testing are explained in the next section and the code
 to reproduce the results is in the attached notebooks.
@@
 
-shows that using paged aggregation alone is not a complete solution. This behavior us caused by over-reads of data now distributed in pages and the internals of HDF5 not knowing how to optimize 
+Using paged aggregation alone is not a complete solution. This behavior is caused by over-reads of data now distributed in pages and the internals of HDF5 not knowing how to optimize 
 the requests.
@@
 
-shows that performance once the I/O configuration is aligned with the chunking in the file, access times perform on par with cloud optimized access patterns like Kerchunk/Zarr. 
-These numbers are from in-region execution. Out of region is considerable slower for the non cloud optimized case.
+Once the I/O configuration is aligned with the chunking in the file, access times perform on par with cloud optimized access patterns like Kerchunk/Zarr. 
+These numbers are from in-region execution. Out of region is considerably slower for the non-cloud-optimized case.
 
@@
 
-Based on the benckmarks we got from our tests, we have split our recommendations for the ATL03 product into 3 main categories, creating the files, accessing the files, and future tool development.
+Based on the benckmarks we got from our tests, we have split our recommendations for the ATL03 product into 3 main categories: creating the files, accessing the files, and future tool development.
 These recommendations aim to streamline HDF5 workflows in cloud environments, enhancing performance and reducing costs.
@@
 Based on our testing we recommend the following cloud optimizations for creating HDF5 files for the ATL03 product:
-Create HDF5 files using paged aggregation by setting HDF5 library parameters:
 
-1. File page strategy: H5F_FSPACE_STRATEGY_PAGE
-2. File page size: 8000000
+1. Create HDF5 files using paged aggregation by setting HDF5 library parameters:
+    a. File page strategy: H5F_FSPACE_STRATEGY_PAGE
+    b. File page size: 8000000
       If repacking an existing file, h5repack contains the code to alter these variables inside the file 
@@
 
-Based on the variable size of ATL03 it becomes really difficult to allocate a fixed metadata page, big files contain north of 30MB of metadata, but the median sized file is below 8MB. If we had
+Based on the variable size of ATL03 it becomes really difficult to allocate a fixed metadata page.
+Big files contain north of 30MB of metadata, but the median metadata size per file is below 8MB.
+If we had
 adopted user block we would have caused an increase in the file size and storage cost of approximate 30% (reference to our tests).
@@
 Avoid small chunks, as they cause excessive HTTP overhead and slower access speeds.
-- **Parallel Access**: Use parallel computing frameworks like `Dask` or multiprocessing to divide read operations across multiple processes or nodes.
+- **Parallel Access**: Use parallel computing frameworks like [`Dask`](https://www.dask.org/) or multiprocessing to divide read operations across multiple processes or nodes.
 This alleviates the sequential access bottleneck caused by the HDF5 global lock, particularly in workflows accessing multiple datasets.
@@
 2. Set the "file space page size" to 8MiB.
-3. Changed all "COMPACT" dataset storage types to "CONTIGUOUS".
-4. Increased the "chunk size" of the photon-rate datasets (from 10,000 to 100,000 elements), while making sure no "chunk sizes" exceeded the 8MiB "file space page size".
-5. Introduced a new production step that executes the "h5repack" utility (with no options) to create a "defragmented" final product.
+3. Change all "COMPACT" dataset storage types to "CONTIGUOUS".
+4. Increase the "chunk size" of the photon-rate datasets (from 10,000 to 100,000 elements), while making sure no "chunk sizes" exceed the 8MiB "file space page size".
+5. Introduce a new production step that executes the "h5repack" utility (with no options) to create a "defragmented" final product.
 

### 2024-12-18 paper.qmd

@@
 This table represents the different configurations we used for our tests in 2 file sizes.
-It's worth noticing we encountered a few outlier cases where compression and chunk sizes led page aggregation to an increase in file size of approximately 10% which was above the desired value for NSIDC (5% max)
+It's worth noticing we encountered a few outlier cases where compression and chunk sizes led page aggregation to an increase in file size of approximately 10% which was above the desired value for NSIDC (5% max).
 We tested these files using the most common libraries to handle HDF5 and 2 different I/O drivers that support remote access to AWS S3, fsspec and the native S3.

### 2024-12-19 _tex/paper.tex
Built site for gh-pages

@@
 to take into account. The first thing to observe is that page
-aggregation will --as we mentioned-- consolidate the file-level metadata
-at the front of the file and will add information in the so-called
-superblock\footnote{The HDF5 superblock is a crucial component of the
-  HDF5 file format, acting as the starting point for accessing all data
-  within the file. It stores important metadata such as the version of
-  the file format, pointers to the root group, and addresses for
-  locating different file components} The next thing to notice is that
-page size us uses across the board for metadata and data as of October
+aggregation will -- as we mentioned -- consolidate the file-level
+metadata at the front of the file and will add information in the
+so-called superblock\footnote{The HDF5 superblock is a crucial component
+  of the HDF5 file format, acting as the starting point for accessing
+  all data within the file. It stores important metadata such as the
+  version of the file format, pointers to the root group, and addresses
+  for locating different file components} The next thing to notice is
+page size is used across the board for metadata and data as of October
 2024 and version 1.14 of the HDF5 library, page size cannot dynamically
@@
 This table represents the different configurations we used for our tests
-in 2 file sizes. It's worth noticing that we encountered a few outlier
-cases where compression and chunk sizes led page aggregation to an
-increase in file size of approximately 10\% which was above the desired
-value for NSIDC (5\% max) We tested these files using the most common
-libraries to handle HDF5 and 2 different I/O drivers that support remote
-access to AWS S3, fsspec and the native S3. The results of our testing
-is explained in the next section and the code to reproduce the results
-is in the attached notebooks.
+in 2 file sizes. It's worth noticing we encountered a few outlier cases
+where compression and chunk sizes led page aggregation to an increase in
+file size of approximately 10\% which was above the desired value for
+NSIDC (5\% max). We tested these files using the most common libraries
+to handle HDF5 and 2 different I/O drivers that support remote access to
+AWS S3, fsspec and the native S3. The results of our testing are
+explained in the next section and the code to reproduce the results is
+in the attached notebooks.
 
@@
 
-\caption{\label{fig-4}shows that using paged aggregation alone is not a
-complete solution. This behavior us caused by over-reads of data now
-distributed in pages and the internals of HDF5 not knowing how to
-optimize the requests. This means that if we cloud optimize alone and
-use the same code, in some cases we'll make access to these files even
-slower. A very important thing to notice here is that rechunking the
-file, in this case using 10X bigger chunks results in a predictable 10X
-improvement in access times without any cloud optimization involved.
-Having less chunks generates less metadata and bigger requests, in
-general is it recommended that chunk sizes should range between 1MB and
-10MB{[}Add citation, S3 and HDF5{]} and if we have enough memory and
-bandwidth even bigger (Pangeo recommends up to 100MB chunks){[}Add
-citation.{]}}
+\caption{\label{fig-4}Using paged aggregation alone is not a complete
+solution. This behavior is caused by over-reads of data now distributed
+in pages and the internals of HDF5 not knowing how to optimize the
+requests. This means that if we cloud optimize alone and use the same
+code, in some cases we'll make access to these files even slower. A very
+important thing to notice here is that rechunking the file, in this case
+using 10X bigger chunks results in a predictable 10X improvement in
+access times without any cloud optimization involved. Having less chunks
+generates less metadata and bigger requests, in general is it
+recommended that chunk sizes should range between 1MB and 10MB{[}Add
+citation, S3 and HDF5{]} and if we have enough memory and bandwidth even
+bigger (Pangeo recommends up to 100MB chunks){[}Add citation.{]}}
 
@@
 
-\caption{\label{fig-5}shows that performance once the I/O configuration
-is aligned with the chunking in the file, access times perform on par
-with cloud optimized access patterns like Kerchunk/Zarr. These numbers
-are from in-region execution. Out of region is considerable slower for
-the non cloud optimized case.}
+\caption{\label{fig-5}Once the I/O configuration is aligned with the
+chunking in the file, access times perform on par with cloud optimized
+access patterns like Kerchunk/Zarr. These numbers are from in-region
+execution. Out of region is considerably slower for the
+non-cloud-optimized case.}
 
@@
 Based on the benckmarks we got from our tests, we have split our
-recommendations for the ATL03 product into 3 main categories, creating
+recommendations for the ATL03 product into 3 main categories: creating
 the files, accessing the files, and future tool development. These
@@
 Based on our testing we recommend the following cloud optimizations for
-creating HDF5 files for the ATL03 product: Create HDF5 files using paged
-aggregation by setting HDF5 library parameters:
+creating HDF5 files for the ATL03 product:
 
@@
 \def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  Create HDF5 files using paged aggregation by setting HDF5 library
+  parameters:
+
+  \begin{enumerate}
+  \def\labelenumii{\alph{enumii}.}
   \tightlist
@@
 \end{Shaded}
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\setcounter{enumi}{2}
-\tightlist
 \item
@@
 Based on the variable size of ATL03 it becomes really difficult to
-allocate a fixed metadata page, big files contain north of 30MB of
-metadata, but the median sized file is below 8MB. If we had adopted user
-block we would have caused an increase in the file size and storage cost
-of approximate 30\% (reference to our tests). Another consequence of
-using a dedicated fixed page for file-level metadata is that metadata
-overflow will generate the same impact in access times, the library will
-fetch the metadata in one go but the rest will be using the predefined
-block size of 4kb.
+allocate a fixed metadata page. Big files contain north of 30MB of
+metadata, but the median metadata size per file is below 8MB. If we had
+adopted user block we would have caused an increase in the file size and
+storage cost of approximate 30\% (reference to our tests). Another
+consequence of using a dedicated fixed page for file-level metadata is
+that metadata overflow will generate the same impact in access times,
+the library will fetch the metadata in one go but the rest will be using
+the predefined block size of 4kb.
 
@@
   \textbf{Parallel Access}: Use parallel computing frameworks like
-  \texttt{Dask} or multiprocessing to divide read operations across
-  multiple processes or nodes. This alleviates the sequential access
-  bottleneck caused by the HDF5 global lock, particularly in workflows
-  accessing multiple datasets.
+  \href{https://www.dask.org/}{\texttt{Dask}} or multiprocessing to
+  divide read operations across multiple processes or nodes. This
+  alleviates the sequential access bottleneck caused by the HDF5 global
+  lock, particularly in workflows accessing multiple datasets.
 \item
@@
 \item
-  Changed all ``COMPACT'' dataset storage types to ``CONTIGUOUS''.
+  Change all ``COMPACT'' dataset storage types to ``CONTIGUOUS''.
 \item
-  Increased the ``chunk size'' of the photon-rate datasets (from 10,000
-  to 100,000 elements), while making sure no ``chunk sizes'' exceeded
-  the 8MiB ``file space page size''.
+  Increase the ``chunk size'' of the photon-rate datasets (from 10,000
+  to 100,000 elements), while making sure no ``chunk sizes'' exceed the
+  8MiB ``file space page size''.
 \item
-  Introduced a new production step that executes the ``h5repack''
-  utility (with no options) to create a ``defragmented'' final product.
+  Introduce a new production step that executes the ``h5repack'' utility
+  (with no options) to create a ``defragmented'' final product.
 \end{enumerate}

### 2025-01-15 paper.qmd
modify text to reflect access protocol

@@
   The Hierarchical Data Format (HDF) is a common archival format for n-dimensional scientific data; it has been utilized to store valuable information from astrophysics to earth sciences and everything in between.
-As flexible and powerful as HDF can be, it comes with big tradeoffs when it’s accessed from remote storage systems, mainly because the file format and the client I/O libraries were designed for local and supercomputing workflows.
+As flexible and powerful as HDF can be, it comes with big tradeoffs when files on remote storage systems are accessed using HTTPS (Hypertext Transfer Protocol Secure), mainly because the file format and the client I/O libraries were designed for local and supercomputing workflows.
 As scientific data and workflows migrate to the cloud , efficient access to data stored in HDF format is a key factor that will accelerate or slow down “science in the cloud” across all disciplines.

### 2025-01-15 paper.qmd
fix typo

@@
 As flexible and powerful as HDF can be, it comes with big tradeoffs when files on remote storage systems are accessed using HTTPS (Hypertext Transfer Protocol Secure), mainly because the file format and the client I/O libraries were designed for local and supercomputing workflows.
-As scientific data and workflows migrate to the cloud , efficient access to data stored in HDF format is a key factor that will accelerate or slow down “science in the cloud” across all disciplines.
+As scientific data and workflows migrate to the cloud, efficient access to data stored in HDF format is a key factor that will accelerate or slow down “science in the cloud” across all disciplines.
   We present an implementation of recently available features in the HDF5 stack that results in performant access to HDF from remote cloud storage.

### 2025-01-15 paper.qmd
efficient access will accelerate science

@@
 As flexible and powerful as HDF can be, it comes with big tradeoffs when files on remote storage systems are accessed using HTTPS (Hypertext Transfer Protocol Secure), mainly because the file format and the client I/O libraries were designed for local and supercomputing workflows.
-As scientific data and workflows migrate to the cloud, efficient access to data stored in HDF format is a key factor that will accelerate or slow down “science in the cloud” across all disciplines.
+As scientific data and workflows migrate to the cloud, efficient access to data stored in HDF format is a key factor that will accelerate “science in the cloud” across all disciplines.
   We present an implementation of recently available features in the HDF5 stack that results in performant access to HDF from remote cloud storage.

### 2025-01-16 paper.qmd
Fix title
add "archived in"
remove unnecessary qualification

@@
 
-## Problem
+## The Problem
 
-Scientific data from NASA and other agencies are increasingly being distributed from the commercial cloud.
+Scientific data from NASA and other agencies are increasingly being archived in and distributed from the commercial cloud.
 Cloud storage enables large-scale workflows and should reduce local storage costs.
-It also allows the use of scalable on-demand cloud computing resources by individual scientists and the broader scientific community.
+It also allows the use of scalable on-demand cloud computing resources.
 However, the majority of this scientific data is stored in a format that was not designed for the cloud: The Hierarchical Data format or HDF.

### 2025-01-16 paper.qmd
rewrite sensence to change to present tense

@@
 
-The most recent version of the Hierarchical Data Format is HDF5, a common archival format for n-dimensional scientific data; it has been utilized to store valuable information from astrophysics to earth sciences and everything in between.
-As flexible and powerful as HDF5 can be, it comes with big trade-offs when it’s accessed from remote storage systems.
+The most recent version of the Hierarchical Data Format is HDF5, a common archival format for n-dimensional scientific data.
+HDF5 is used to store data from astrophysics to earth sciences and everything in between.
+However, as flexible and powerful as HDF5 can be, it comes with big trade-offs when it’s accessed from remote storage systems.
 

### 2025-01-16 paper.qmd
name problems after colon

@@
 HDF5 is a complex file format; we can think of it as a file system using a tree-like structure with multiple data types and native data structures.
-Because of this complexity, the most reliable way of accessing data stored in this format is using the HDF5 C API.
-Regardless of access pattern, nearly all tools ultimately rely on the HDF5-C library and this brings a couple issues that affect the efficiency of accessing this format over the network:
+Because of this complexity, the most reliable way of accessing data stored in this format is using the HDF5 C API^[Application Programming Interface].
+Regardless of access pattern, nearly all tools ultimately rely on the HDF5-C library and this brings a couple issues that affect the efficiency of accessing this format over the network: metadata fragmentation and global API lock.
 

### 2025-01-16 paper.qmd
add In HDF5 for clarity
add specific time frame.  Check if correct.

@@
 When working with large datasets, especially those that include numerous variables and nested groups, the storage of file-level metadata can become a challenge.
-By default, metadata associated with each dataset is stored in chunks of 4 kilobytes (KB).
-This chunking mechanism was originally intended to optimize storage efficiency and access speed on disks with hardware resources available more than 20 years ago.
+In HDF5, by default, metadata associated with each dataset is stored in chunks of 4 kilobytes (KB).
+This chunking mechanism was originally intended to optimize storage efficiency and access speed on disks with hardware resources available when HDF was originally designed in the 1990s.
 In datasets with many variables and/or complex hierarchical structures, these 4KB chunks can lead to significant fragmentation.

### 2025-01-16 paper.qmd
suggest rephrasing
add impact of cloud reads

@@
 Fragmentation occurs when this metadata is spread out across multiple non-contiguous chunks within the file.
-This results in inefficiencies when accessing or modifying data because compatible libraries need to read from multiple, scattered locations in the file.
+This makes accessing or modifying data inefficient because the HDF library has to read from multiple, scattered locations in the file.
 Over time, as the dataset grows and evolves, this fragmentation can compound, leading to degraded performance and increased storage overhead.
 In particular, operations that involve reading or writing metadata, such as opening a file, checking attributes, or modifying variables, can become slower and more resource-intensive.
+When cloud-hosted files are accessed over a network using HTTPS, multiple reads increases the time to access data.
 

### 2025-01-16 paper.qmd
change imperative

@@
 Because of the historical complexity of operations with the HDF5 format[@The_HDF_Group_Hierarchical_Data_Format], there has been a necessity to make the library thread-safe and similarly to what happens in the Python language, the simplest mechanism to implement this is to have a global API lock.
-This global lock is not as big of an issue when we read data from local disk but it becomes a major bottleneck when we read data over the network because each read is sequential and latency in the cloud is exponentially bigger than local access [@Mozilla-latency-2024] [@scott-2020].
+This global lock is not as big of an issue when we read data from local disk but it becomes a major bottleneck when we read data over the network because each read has to be sequential and latency in the cloud is exponentially bigger than local access [@Mozilla-latency-2024] [@scott-2020].
 

### 2025-01-16 paper.qmd
modify figure caption

@@
 
-shows how reads (Rn) are done in order to access file metadata, In the first read, R0, the HDF5 library verifies the file signature from the superblock,  subsequent reads, R1, R2,...Rn, read file metadata, 4kb at the time.
+A cartoon showing the distribution of fragmented file-level metadata and data in an HDF5 file.
+In the first read, R0, the HDF5 library verifies the file signature from the superblock.
+Subsequent reads, R1, R2,...Rn, read file-level metadata, 4kb at the time.
+The information in these metadata reads allow the actual data stored in the file to be accessed.
 

### 2025-01-16 paper.qmd
modify headings to fit paper structure

@@
 
-#### **Metadata fragmentation**
+##### _Metadata fragmentation_
 
@@
 Fragmentation occurs when this metadata is spread out across multiple non-contiguous chunks within the file.
-This makes accessing or modifying data inefficient because the HDF library has to read from multiple, scattered locations in the file.
+This makes accessing or modifying data inefficient because the HDF library has to read from multiple, scattered locations in the file (@fig-1).
 Over time, as the dataset grows and evolves, this fragmentation can compound, leading to degraded performance and increased storage overhead.
@@
 
-#### **Global API Lock**
+##### _Global API Lock_
 
-Because of the historical complexity of operations with the HDF5 format[@The_HDF_Group_Hierarchical_Data_Format], there has been a necessity to make the library thread-safe and similarly to what happens in the Python language, the simplest mechanism to implement this is to have a global API lock.
+Because of the historical complexity of operations with the HDF5 format [@The_HDF_Group_Hierarchical_Data_Format], there has been a necessity to make the library thread-safe and similarly to what happens in the Python language, the simplest mechanism to implement this is to have a global API lock.
 This global lock is not as big of an issue when we read data from local disk but it becomes a major bottleneck when we read data over the network because each read has to be sequential and latency in the cloud is exponentially bigger than local access [@Mozilla-latency-2024] [@scott-2020].

### 2025-01-16 paper.qmd
reorganize background, move text to methodology, add text on IS2

@@
 
-#### **Background and data selection**
+Products from NASA's Ice, Cloud, and Land Elevation Satellite 2 (ICESat-2) mission were some of the first NASA datasets to be migrated from on-premises storage at the National Snow and Ice Data Center Distributed Active Archive Center (NSIDC-DAAC) to Amazon Web Services (AWS) Simple Storage Service (S3).
+The ICESat-2 observatory carries the Advanced Topographic Laser Altimeter System (ATLAS), a photon-counting lidar altimeter that measures the round-trip travel time of low-energy green laser light to retrieve the heights of ice sheets, sea ice, land surfaces, vegetation, oceans, and clouds [@NEUMANN2019111325].
+The geolocated photon height product (ATL03), which is the basic science product from which other higher-level, surface-specific, products are derived, retrieves the latitude, longitude and ellipsoid heights of photons every 70 cm along six ground tracks as the satellite orbits the Earth.
+All ICESat-2 data products are in HDF5 format.
+The ICESat-2 mission produces # TerraBytes of data each day.
+ATL03 files range between # GB and # TB, with an average file size # GB.
+Other surface-specific products tend to have smaller file sizes.
+Individual file sizes and the volume of data required for most science goals make traditional download-and-analyse workflows time consuming and require local storage.
+However, cloud-based workflows, in which ICESat-2 file objects in AWS S3 buckets are accessed from cloud compute instances in the same AWS region as the data, cannot take full advantage of direct download and scaleability offered by these workflows because of the large latency caused by metadata fragmentation and global api lock.
 
-As a result of community feedback and “hack weeks” organized by NSIDC and UW eScience Institute in 2023[@h5cloud2023], NSIDC started the Cloud Optimized Format Investigation (COFI) project to improve access to HDF5 from the ICESat-2 mission, a spaceborne lidar that retrieves surface topography of the Earth’s ice sheets, land and oceans [@NEUMANN2019111325].
-Because of its complexity, large size and importance for cryospheric studies we targeted the ATL03 data product.
-The most relevant variable in ATL03 are geolocated photon heights from the ICESat-2 ATLAS instrument.
-Each ATL03 file contains 1003 geophysical variables in 6 data groups.
-Although our research was focused on this dataset, most of our findings are applicable to any dataset stored in HDF5 and NetCDF4.
+NSIDC started the Cloud Optimized Format Investigation (COFI) project to improve access to HDF5 from the ICESat-2 mission (and other missions) in response to feedback from the ICESat-2 community and the experiences of particpants attending ICESat-2 hack weeks, organized University of Washington eScience Institute.
+The initial effort on the COFI project was made at the 2023 hack week by a collection of developers from NASA DAACs, private companies, and scientists and students [@h5cloud2023].
+This collaborative effort has continued led by NSIDC with the HDF Group and NASA engineers.
 
@@
 
+We have targeted the ATL03, geolocated photon height, product because of its complexity, and large size, and because this data is the input to all standard surface-specific data products generated by NASA, as well as the starting point for research and analysis using ICESat-2 products.
+ATL03 HDF5 files contain 1003 variables^[In HDF5, variables are called datasets] organized in # groups^[Define groups here].
+The most relevant variable in ATL03 are geolocated photon heights from the ICESat-2 ATLAS instrument.
+Although our research was focused on this dataset, most of our findings are applicable to any dataset stored in HDF5 and NetCDF4.
+
 We tested access times to original and different configurations of cloud-optimized HDF5 [ATL03 files](https://its-live-data.s3.amazonaws.com/index.html#test-space/cloud-experiments/h5cloud/) stored in AWS S3 buckets in region us-west-2, the region hosting NASA’s Earthdata Cloud archives.
-Files were accessed using Python tools commonly used by Earth scientists: h5py and Xarray[@Hoyer2017-su].
+Files were accessed using Python tools commonly used by Earth scientists: h5py and xarray[@Hoyer2017-su].
 h5py is a Python wrapper around the HDF5 C API.

### 2025-02-16 paper.qmd

@@
 abstract: |
-  The Hierarchical Data Format (HDF) is a common archival format for n-dimensional scientific data; it has been utilized to store valuable information from astrophysics to earth sciences and everything in between.
-As flexible and powerful as HDF can be, it comes with big tradeoffs when files on remote storage systems are accessed using HTTPS (Hypertext Transfer Protocol Secure), mainly because the file format and the client I/O libraries were designed for local and supercomputing workflows.
-As scientific data and workflows migrate to the cloud, efficient access to data stored in HDF format is a key factor that will accelerate “science in the cloud” across all disciplines.
-  We present an implementation of recently available features in the HDF5 stack that results in performant access to HDF from remote cloud storage.
+  The Hierarchical Data Format Version 5 (HDF5) is a common archival format for n-dimensional scientific data; it has been utilized to store valuable information from astrophysics to earth sciences and everything in between.
+As flexible and powerful as HDF5 can be, it comes with big tradeoffs when files on remote storage systems are accessed using HTTPS (Hypertext Transfer Protocol Secure), mainly because the file format and the client I/O libraries were designed for local and supercomputing workflows.
+As scientific data and workflows migrate to the cloud, efficient access to data stored in HDF5 format is a key factor that will accelerate “science in the cloud” across all disciplines.
+  We present an implementation of recently available features in the HDF5 stack that results in performant access to HDF5 files in remote cloud storage.
 This performance is on par with modern cloud-native formats like Zarr but with the advantage of not having to reformat data or generate metadata sidecar files (DMR++, Kerchunk).
@@
 It also allows the use of scalable on-demand cloud computing resources.
-However, the majority of this scientific data is stored in a format that was not designed for the cloud: The Hierarchical Data format or HDF.
+However, the majority of this scientific data are stored in a format that was not designed for the cloud: The Hierarchical Data Format Version 5 (HDF5).
 
-The most recent version of the Hierarchical Data Format is HDF5, a common archival format for n-dimensional scientific data.
-HDF5 is used to store data from astrophysics to earth sciences and everything in between.
-However, as flexible and powerful as HDF5 can be, it comes with big trade-offs when it’s accessed from remote storage systems.
+HDF5 is a generic n-dimensional array format and have been used to store scientific data from astrophysics to earth sciences and everything in between.
+However, as flexible and powerful as HDF5 can be, it comes with big trade-offs when it's accessed from remote storage systems.
 
-HDF5 is a complex file format; we can think of it as a file system using a tree-like structure with multiple data types and native data structures.
-Because of this complexity, the most reliable way of accessing data stored in this format is using the HDF5 C API^[Application Programming Interface].
-Regardless of access pattern, nearly all tools ultimately rely on the HDF5-C library and this brings a couple issues that affect the efficiency of accessing this format over the network: metadata fragmentation and global API lock.
+HDF5 is a complex file format; we can think of it as a file system implementing a tree-like structure of n-dimensional arrays of many different data types.
+Because of this complexity, the most reliable way of accessing data stored in this format is using the HDF5 library^[Application Programming Interface].
+Regardless of access pattern, nearly all software ultimately rely on the library and this brings a couple of issues that affect the efficiency of accessing this format over the network: metadata fragmentation and global API lock.
 
@@
 
-When working with large datasets, especially those that include numerous variables and nested groups, the storage of file-level metadata can become a challenge.
-In HDF5, by default, metadata associated with each dataset is stored in chunks of 4 kilobytes (KB).
-This chunking mechanism was originally intended to optimize storage efficiency and access speed on disks with hardware resources available when HDF was originally designed in the 1990s.
-In datasets with many variables and/or complex hierarchical structures, these 4KB chunks can lead to significant fragmentation.
-
-Fragmentation occurs when this metadata is spread out across multiple non-contiguous chunks within the file.
-This makes accessing or modifying data inefficient because the HDF library has to read from multiple, scattered locations in the file (@fig-1).
-Over time, as the dataset grows and evolves, this fragmentation can compound, leading to degraded performance and increased storage overhead.
+When working with large datasets, especially those that include numerous variables and nested groups, the storage of file content-related (internal) metadata can become a challenge.
+In HDF5, by default, internal metadata associated with each HDF5 dataset^[In HDF5, variables are called datasets] is stored in blocks of 4 kilobytes (KB).
+This approach was originally intended to optimize storage efficiency, worklow flexibility, and access speed on disks with hardware resources back when HDF5 was initially designed in the 1990s.
+In datasets with many variables and/or complex hierarchical structures, there could be many such 4KB blocks dispersed throughout the file which causes significant fragmentation of internal metadata information.
+This makes accessing HDF5 files in cloud object stores inefficient because HDF5 library has to read over a network using HTTPS from multiple, scattered locations in a file (@fig-1), which increases the time to access data significantly.
+**Over time, as the dataset grows and evolves, this fragmentation can compound, leading to degraded performance and increased storage overhead.
 In particular, operations that involve reading or writing metadata, such as opening a file, checking attributes, or modifying variables, can become slower and more resource-intensive.
-When cloud-hosted files are accessed over a network using HTTPS, multiple reads increases the time to access data.
+(???)**
 
@@
 
-Because of the historical complexity of operations with the HDF5 format [@The_HDF_Group_Hierarchical_Data_Format], there has been a necessity to make the library thread-safe and similarly to what happens in the Python language, the simplest mechanism to implement this is to have a global API lock.
+Because of the need to support complex data I/O operations with the HDF5 format [@The_HDF_Group_Hierarchical_Data_Format], there has been a necessity to make the library thread-safe, and similarly to what happens in the Python interpreter, the simplest mechanism to implement this is to have a global API lock.
 This global lock is not as big of an issue when we read data from local disk but it becomes a major bottleneck when we read data over the network because each read has to be sequential and latency in the cloud is exponentially bigger than local access [@Mozilla-latency-2024] [@scott-2020].
@@
 In the first read, R0, the HDF5 library verifies the file signature from the superblock.
-Subsequent reads, R1, R2,...Rn, read file-level metadata, 4kb at the time.
+Subsequent reads, R1, R2,...Rn, read internal metadata, 4KB at the time.
 The information in these metadata reads allow the actual data stored in the file to be accessed.
@@
 The geolocated photon height product (ATL03), which is the basic science product from which other higher-level, surface-specific, products are derived, retrieves the latitude, longitude and ellipsoid heights of photons every 70 cm along six ground tracks as the satellite orbits the Earth.
-All ICESat-2 data products are in HDF5 format.
+All ICESat-2 data products are in the HDF5 format.
 The ICESat-2 mission produces # TerraBytes of data each day.
@@
 Individual file sizes and the volume of data required for most science goals make traditional download-and-analyse workflows time consuming and require local storage.
-However, cloud-based workflows, in which ICESat-2 file objects in AWS S3 buckets are accessed from cloud compute instances in the same AWS region as the data, cannot take full advantage of direct download and scaleability offered by these workflows because of the large latency caused by metadata fragmentation and global api lock.
+However, cloud-based workflows, in which ICESat-2 file objects in AWS S3 buckets are accessed from cloud compute instances in the same AWS region as the data, cannot take full advantage of direct download and scaleability offered by these workflows because of the large latency caused by metadata fragmentation and global API lock.
 
-NSIDC started the Cloud Optimized Format Investigation (COFI) project to improve access to HDF5 from the ICESat-2 mission (and other missions) in response to feedback from the ICESat-2 community and the experiences of particpants attending ICESat-2 hack weeks, organized University of Washington eScience Institute.
+NSIDC started the Cloud Optimized Format Investigation (COFI) project to improve access to HDF5 from the ICESat-2 mission (and other missions) in response to feedback from the ICESat-2 community and the experiences of particpants attending ICESat-2 hack weeks, organized  by the University of Washington eScience Institute.
 The initial effort on the COFI project was made at the 2023 hack week by a collection of developers from NASA DAACs, private companies, and scientists and students [@h5cloud2023].
@@
 
-We have targeted the ATL03, geolocated photon height, product because of its complexity, and large size, and because this data is the input to all standard surface-specific data products generated by NASA, as well as the starting point for research and analysis using ICESat-2 products.
-ATL03 HDF5 files contain 1003 variables^[In HDF5, variables are called datasets] organized in # groups^[Define groups here].
+We have targeted the ATL03, geolocated photon height, product because of its complexity, and large size, and because these data are the input to all standard surface-specific scientific products generated by NASA, as well as the starting point for research and analysis using ICESat-2 products.
+ATL03 HDF5 files contain 1003 variables organized in # groups^[Define groups here.
+Actually, the term has been used earlier in the text.].
 The most relevant variable in ATL03 are geolocated photon heights from the ICESat-2 ATLAS instrument.
@@
 
-We tested access times to original and different configurations of cloud-optimized HDF5 [ATL03 files](https://its-live-data.s3.amazonaws.com/index.html#test-space/cloud-experiments/h5cloud/) stored in AWS S3 buckets in region us-west-2, the region hosting NASA’s Earthdata Cloud archives.
+We tested access times to original and different configurations of cloud-optimized HDF5 [ATL03 files](https://its-live-data.s3.amazonaws.com/index.html#test-space/cloud-experiments/h5cloud/) stored in AWS S3 buckets in region `us-west-2`, the region hosting NASA’s Earthdata Cloud archives.
 Files were accessed using Python tools commonly used by Earth scientists: h5py and xarray[@Hoyer2017-su].
-h5py is a Python wrapper around the HDF5 C API.
+h5py is a Python wrapper around the HDF5 library.
 xarray^[`h5py` is a dependency of Xarray] is a widely used Python package for working with n-dimensional data.
-We also tested access times using h5coro, a python package optimized for reading HDF5 files from S3 buckets and kerchunk, a tool that creates an efficient lookup table for file chunks to allow performant partial reads of files.
+We also tested access times using h5coro, a Python package optimized for reading HDF5 files from S3 buckets, and kerchunk, a tool that creates an efficient lookup table for file chunks to allow performant partial reads of HDF5 files without using the library.
 
-The test files were originally cloud optimized by  “repacking” them, using a relatively new feature in the HDF5 C API called “paged aggregation”.
-Page aggregation does 2 things: first, it collects file-level metadata from datasets and stores it on dedicated metadata blocks at the front of the file; second, it forces the library to write both data and metadata using these fixed-size pages.
-Aggregation allows client libraries to read  file metadata with only a few requests using the page size as a fixed request size, overriding the 1 request per chunk behavior.
+The test files were originally cloud optimized by “repacking” them, using a relatively recent HDF5 feature called _paged aggregation file space management strategy_.
+This feature instructs the library to write out both data and internal metadata into separate _file pages_ of a specific size set at the file's creation.
+By storing previously small and scattered internal metadata blocks into a bigger _metadata file page_, this vital information for data access is effectively consolidated into larger blocks.
+Such blocks are much more suitable for slow HTTPS-based access of cloud object stores.
+With an appropriately sized file page, all internal file metadata can fit into one file page, which will be located at the start of the file.
+
+HDF5 library read operations are always one file page for page aggregated files, which means that, in an ideal case, the entire information about what is and where in a file can be obtained with just one HTTPS request.
+Furthermore, with file's actual data now stored in _raw data file pages_, the library can read several variables with one request as long as they are in the same file page.
 
@@
 
-shows how file-level metadata and data gets internally packed once we use paged aggregation on a file.
+shows how internal metadata and data get laid out in case of a page aggregated file.
 
@@
 As we can see in [@fig-2], when we cloud optimize a file using paged-aggregation there are some considerations and behavior that we had to take into account. The first thing to observe is that
-page aggregation will -- as we mentioned -- consolidate the file-level metadata at the front of the file and will add information in the so-called superblock^[The HDF5 superblock is a crucial component of the HDF5 file format, acting as the starting point for accessing all data within the file.
-It stores important metadata such as the version of the file format, pointers to the root group, and addresses for locating different file components]
-The next thing to notice is page size is used across the board for metadata and data as of October 2024 and version 1.14 of the HDF5 library, page size cannot dynamically adjust to the total metadata size.
+page aggregation will -- as we mentioned -- consolidate the internal metadata at the front of the file and will add information in the so-called superblock^[The HDF5 superblock is a crucial component of the HDF5 file format, acting as the starting point for accessing all data within the file.
+It stores important metadata such as the version of the file format, pointers to the root group, and addresses for locating different file components.]
+The next thing to notice is the same file page size is used for both metadata and raw data pages.
+Page size cannot dynamically adjust to the total metadata size because the total size is not known prior to all the data are stored in the file.
 
@@
 
-shows how file-level metadata and data packing inside aggregated pages leave unused space that can potentially increase the file size in a considerable way.
+shows how internal metadata and data packing inside file pages leave unused space that can potentially increase the file size in a considerable way.
 
@@
 
-This one page size for all approach simplifies how the HDF5 API reads the file (if configured) but it also brings unused page space and chunk over-reads.
-In the case of the ICESat-2 dataset (ATL03) the data itself has been partitioned and each granule
-represents a segment in the satellite orbit and within the file the most relevant dataset is chunked using 10,000 items per chunk, with data being float-32 and using a fast compression value, the resulting chunk size is on average under 40KB, which is really small
+This one page size for all approach simplifies how the HDF5 library reads the file (if configured) but it also brings unused page space and chunk over-reads.
+In the case of the ICESat-2 dataset (ATL03) the data itself have been partitioned and each granule
+represents a segment in the satellite orbit and within the file the most relevant HDF5 dataset is chunked using 10,000 array elements per chunk, with data being float-32 and using a fast compression value, the resulting chunk size is on average under 40KB, which is really small
 for an HTTP request, especially when we have to read them sequentially.
-Because of these considerations, we opted for testing different page sizes, and increase chunk size.
+Because of these considerations, we opted for testing different file page sizes and increase chunk size.
 The following table describes the different configurations used in our tests.
@@
 |------------------------	|----------------------------------------------------	|----------------------	|---------------	|-----------	|-----------	|----------------	|
-| original               	| original file from ATL03 v6 (1gb and 7gb)          	| 0                    	| 1.5km         	| (10000,)  	| N/A       	| 35kb           	|
-| original-kerchunk      	| kerchunk sidecar of the original file              	| N/A                  	| 1.5km         	| (10000,)  	| N/A       	| 35kb           	|
-| page-only-4mb          	| paged-aggregated file with 4mb per page            	| ~1%                  	| 1.5km         	| (10000,)  	| 4MB       	| 35kb           	|
-| page-only-8mb          	| paged-aggregated file with 4mb per pag8            	| ~1%                  	| 1.5km         	| (10000,)  	| 8MB       	| 35kb           	|
-| rechunked-4mb          	| page-aggregated and bigger chunk sizes             	| ~1%                  	| 10km          	| (100000,) 	| 4MB       	| 400kb          	|
-| rechunked-8mb          	| page-aggregated and bigger chunk sizes             	| ~1%                  	| 10km          	| (100000,) 	| 8MB       	| 400kb          	|
-| rechunked-8mb-kerchunk 	| kerchunk sidecar of the last paged-aggregated file 	| N/A                  	| 10km          	| (100000,)  	| 8MB       	| 400kb          	|
+| original               	| original file from ATL03 v6 (1GB and 7GB)          	| 0                    	| 1.5km         	| (10000,)  	| N/A       	| 35KB           	|
+| original-kerchunk      	| kerchunk sidecar of the original file              	| N/A                  	| 1.5km         	| (10000,)  	| N/A       	| 35KB           	|
+| page-only-4mb          	| paged-aggregated file with 4MB per page            	| ~1%                  	| 1.5km         	| (10000,)  	| 4MB       	| 35KB           	|
+| page-only-8mb          	| paged-aggregated file with 4MB per page            	| ~1%                  	| 1.5km         	| (10000,)  	| 8MB       	| 35KB           	|
+| rechunked-4mb          	| page-aggregated and bigger chunk sizes             	| ~1%                  	| 10km          	| (100000,) 	| 4MB       	| 400KB          	|
+| rechunked-8mb          	| page-aggregated and bigger chunk sizes             	| ~1%                  	| 10km          	| (100000,) 	| 8MB       	| 400KB          	|
+| rechunked-8mb-kerchunk 	| kerchunk sidecar of the last paged-aggregated file 	| N/A                  	| 10km          	| (100000,)  	| 8MB       	| 400KB          	|
 
@@
 It's worth noticing we encountered a few outlier cases where compression and chunk sizes led page aggregation to an increase in file size of approximately 10% which was above the desired value for NSIDC (5% max).
-We tested these files using the most common libraries to handle HDF5 and 2 different I/O drivers that support remote access to AWS S3, fsspec and the native S3.
-The results of our testing are explained in the next section and the code
-to reproduce the results is in the attached notebooks.
+We tested these files using the most common libraries to handle HDF5 and 2 different I/O drivers that support remote access to AWS S3, Python's fsspec package and the library's native Read-Only S3 (ROS3) driver.
+The results of our testing are explained in the next section and the code to reproduce the results is in the attached notebooks.
 
@@
 A very important thing to notice here is that rechunking the file, in this case using 10X bigger chunks results in a predictable 10X improvement in access times without any cloud optimization involved.
-Having less chunks generates less metadata and bigger requests, in general is it recommended that chunk sizes should range between 1MB and 10MB[Add citation, S3 and HDF5] and if we have enough memory and bandwidth even
+Having less chunks generates less internal metadata and bigger requests, in general is it recommended that chunk sizes should range between 1MB and 10MB[Add citation, S3 and HDF5] and if we have enough memory and bandwidth even
 bigger (Pangeo recommends up to 100MB chunks)[Add citation.]
@@
 1. Create HDF5 files using paged aggregation by setting HDF5 library parameters:
-    a. File page strategy: H5F_FSPACE_STRATEGY_PAGE
-    b. File page size: 8000000
-      If repacking an existing file, h5repack contains the code to alter these variables inside the file 
+    a. File page strategy: `H5F_FSPACE_STRATEGY_PAGE`
+    b. File page size: 8000000 bytes
+      If repacking an existing file, the `h5repack` command tool contains the code to alter these variables inside the file:
       ```bash
@@
       ```
-3. Avoid using unlimited dimensions when creating variables because the HDF5 API cannot support it inside buffered pages and representation of these variables is not supported by Kerchunk.
+2. Avoid using unlimited dimensions when creating variables because the HDF5 API cannot support it inside buffered pages and representation of these variables is not supported by Kerchunk.
 
@@
 
-Based on the variable size of ATL03 it becomes really difficult to allocate a fixed metadata page.
-Big files contain north of 30MB of metadata, but the median metadata size per file is below 8MB.
+Based on the variable size of ATL03 it becomes really difficult to allocate a fixed metadata page size.
+Big files contain north of 30MB of internal metadata, but the median size per file is below 8MB.
 If we had
 adopted user block we would have caused an increase in the file size and storage cost of approximate 30% (reference to our tests).
-Another consequence of using a dedicated fixed page for file-level metadata is
-that metadata overflow will generate the same impact in access times, the library will fetch the metadata in one go but the rest will be using the predefined block size of 4kb.
+Another consequence of using a dedicated fixed page for internal metadata is
+that metadata overflow will generate the same impact in access times, the library will fetch the metadata in one go but the rest will be using the predefined block size of 4KB.
 
-Paged aggregation is thus the simplest way of cloud optimizing an HDF5 file as the metadata will keep filling dedicated pages until all the file-level metadata is stored at the front of the file.
-Chunk sizes cannot be larger than the page size and when chunk sizes are smaller we need to take into account how these chunks will fit on a page, in an ideal scenario all the space
-will be filled but that is not the case and we will end up with unused space [See @fig-2].
+Paged aggregation is thus the simplest way of cloud optimizing an HDF5 file as the internal metadata will keep filling as many dedicated file pages as required.
+Note that any additional metadata page will not typically be located at the front of the file as there could be some raw data pages written out in the meantime.
+Chunk sizes should be chosen to fit inside a file page, preferably several chunks.
+Any chunk bigger than a page is stored outside of file pages and will require individual S3 request to read it.
+Smaller chunk sizes allow efficient packing by fitting more of them inside a file page and thus reducing its unused space.
+In an ideal scenario all the space of a file page
+will be filled but this is not typically the case [See @fig-2].
 
@@
 However, these recommendations should be applicable across programming languages.
-It's also worth mentioning that the HDF Group aims to include some of these features in their roadmap.
+It's also worth mentioning that the HDF Group aims to include some of these features in the future library releases.
 
@@
 - **Cache Management**: Implement caching for metadata to avoid repetitive fetches.
-Tools like `fsspec` or `h5coro` allow in-memory or on-disk caching for frequently accessed data, reducing latency during high-frequency
+Tools like `fsspec` or `h5coro` allow in-memory or on-disk caching for frequently accessed data, reducing latency during high-frequency.
+HDF5 library has a special cache, called _page buffer_, for caching file pages.
+When combined with the dataset chunk cache, the need to make repeated requests for file's data may be significantly reduced, thus, immediately providing much better performance.
 - **Regional Access**: Operate workflows in the same cloud region as the data to minimize costs and latency.
@@
 
-1. Set the "file space strategy" to H5F_FSPACE_STRATEGY_PAGE and enabled "free space tracking" within the HDF5 file creation property list.
+1. Set the "file space strategy" to `H5F_FSPACE_STRATEGY_PAGE` and enable "free space tracking" within the HDF5 file creation property list.
 2. Set the "file space page size" to 8MiB.
-3. Change all "COMPACT" dataset storage types to "CONTIGUOUS".
-4. Increase the "chunk size" of the photon-rate datasets (from 10,000 to 100,000 elements), while making sure no "chunk sizes" exceed the 8MiB "file space page size".
-5. Introduce a new production step that executes the "h5repack" utility (with no options) to create a "defragmented" final product.
+3. Change all "COMPACT" HDF5 dataset storage layout to "CONTIGUOUS".
+4. Increase the chunk shape of the photon-rate datasets (from 10,000 to 100,000 elements), while making sure no chunk sizes exceed the 8MiB file page size.
+5. Introduce a new production step that executes the `h5repack` command without any option to create a defragmented final HDF5 file output.
 
@@
 Finally, advancing ecosystem interoperability involves setting integration standards and aligning with emerging trends such as serverless and edge computing.
-These efforts, combined with community collaboration, will modernize HDF5 to meet the challenges of evolving data-intensive applications.
+These efforts, combined with community collaboration, will modernize HDF5 to meet the challenges of evolving data-intensive cloud-native applications.
 
@@
 
-- **Chunk Shape:** Align chunk dimensions with anticipated access patterns. For example, row-oriented queries benefit from row-aligned chunks. 
+- **Chunk Shape:** Align chunk dimensions with anticipated access patterns. For example, row-oriented queries benefit from row-aligned chunk shapes.
 - **Chunk Size:** Use chunk sizes between 1–10 MB to match cloud storage block sizes.
@@
 
-Finally, we recognize that this study has not been as extensive as it could have been (cross language, multiple datasets) and yet we think we ran into the key scenarios data producers will face when they start producing cloud optimized HDf5 files.
+Finally, we recognize that this study has not been as extensive as it could have been (cross language, multiple datasets) and yet we think we ran into the key scenarios data producers will face when they start producing cloud optimized HDF5 files.
 We think that there is room for improvement and experimentation with various configurations based on real-world scenarios is crucial to determine the best performance.
